---
title: "Projecting Ampliseq WECARE w/o Danish on 1KGP"
author: "Alexey Larionov"
date: "08 May 2022"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/reanalysis_2022/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_relatedness_and_pca")
scripts_folder <- file.path(base_folder,"scripts","s07_relatedness_and_pca")
setwd(scripts_folder)

library(bigsnpr) # for bed_autoSVD() and bed()
library(bigutilsr) # for prob_dist() and tukey_mc_up() for outlier detection
library(hexbin) # for plotting svd loadings
library(ggplot2)
library(dplyr)

NCORES <- 1
#NCORES <- nb_cores() # 2

```

# Read wecare bed file

```{r}

# Location of bed file
bed_file <- file.path(data_folder,"s02_bed_bim_fam","common_biallelic_autosomal_snps_in_HWE.bed")

# Attach PLINK data to R environment
wecare.bed <- bed(bed_file) # bigsnpr::bed

# Explore wecare.bed
wecare.bed
#names(wecare.bed)
#str(wecare.bed)
#attributes(wecare.bed)
#wecare.bed$bedfile
#wecare.bed$address

# Clean-up
rm(bed_file)

```

# Read wecare phenotypes

```{r}

wecare_phenotypes.df <- wecare.bed$fam
dim(wecare_phenotypes.df)
head(wecare_phenotypes.df)

#potential_outliers <- wecare_phenotypes.df$sample.ID %in% c("P6_D05", "P5_E09")
#sum(potential_outliers)

#data_file <- file.path(data_folder,"s04_phenotypes_with_PCs.RData")
#load(data_file)
#dim(updated_phenotypes.df)
#rm(data_file)

```

# Read b37 1KGP data

```{r}

# Location of bed file
project_folder="/Users/alexey/Documents/wecare/reanalysis_2022"
bed_file <- file.path(project_folder,"b37_1kgp","kgp_b37_biallelic_snps_autosomal_common_non_related_no_outliers.bed")

# Read (=attach?) data
kgp.bed <- bed(bed_file) # bigsnpr::bed
kgp.bed

str(kgp.bed)

kgp_samples <- kgp.bed$fam$sample.ID

# Clean-up
rm(bed_file)

```

# Load population descriptions for 1KGP

```{r}

data_file <- file.path(project_folder,"b37_1kgp","igsr-1000_genomes_phase_3_release.tsv")

kgp_pop.df <- read.table(data_file, header=T, sep = "\t")
dim(kgp_pop.df)
rownames(kgp_pop.df) <- kgp_pop.df$Sample.name

kgp_pops <- kgp_pop.df[kgp_samples,"Superpopulation.code"]
table(kgp_pops, useNA = "always")

kgp_pops_colours <- kgp_pops
"blue" -> kgp_pops_colours[kgp_pops == "EUR"]
"darkgrey" -> kgp_pops_colours[kgp_pops == "AFR"]
"pink" -> kgp_pops_colours[kgp_pops == "EAS"]
"yellow" -> kgp_pops_colours[kgp_pops == "SAS"]
"orange" -> kgp_pops_colours[kgp_pops == "AMR"]
table(kgp_pops_colours, useNA = "always")

rm(data_file)

```

# Check variants overlap between 1kgp and wecare

```{r}

wecare_map.df <- wecare.bed$map
dim(wecare_map.df)

kgp_map.df <- kgp.bed$map
dim(kgp_map.df)

x <- inner_join(kgp_map.df, wecare_map.df, by=c("chromosome","physical.pos"))
y <- inner_join(kgp_map.df, wecare_map.df, by=c("chromosome","physical.pos","allele1","allele2"))

dim(x)
head(x)

dim(y)
head(y)

rm(x,y)

```

# Project Wecare to 1KGP

A simple projecting procedure would be multiplying the genotypes by the corresponding PC loadings.  
However, the augmented algorithm (OADP) is used to avoid shrinkage (the shrinkage is becoming evident in PCs > 5).

```{r}

# Calculate the "projection"
proj <- bed_projectPCA(kgp.bed, wecare.bed, ncores = NCORES, roll.size=0)

```

Because of the low number of variants, 
I had to suppress smoothing log-p-values
to pass this point.  

Other way it crashed at this point with the following error message:  

[Step 1/3] Matching variants of reference with target data..
5,359,025 variants to be matched.
20 ambiguous SNPs have been removed.
175 variants have been matched; 0 were flipped and 32 were reversed.

[Step 2/3] Computing (auto) SVD of reference..

Phase of clumping (on MAC) at r^2 > 0.2.. keep 118 variants.
Discarding 0 variant with MAC < 10.

Iteration 1:
Computing SVD..
Error in bigutilsr::rollmean(S.col[ind], roll.size) : 
  Parameter 'size' is too large.

This error was not yet resolved as on 04Mar2021:  
https://github.com/privefl/bigutilsr/issues/2  

The same error was also happening in the script  
s04_calculate_PCs.Rmd  

```{r}
# Explore proj ...
str(proj)
# proj$obj.svd.ref # big_SVD object computed from reference data.

```

# Plot 2 top PC-s

```{r}

# Get 1KGP PCs
kgen_pcs.mx <- predict(proj$obj.svd.ref)
dim(kgen_pcs.mx)

# Get wecare projections
wecare_pcs.mx <- proj$OADP_proj 
dim(wecare_pcs.mx)
# OADP projection of new data into space of reference PCA
# proj$obj.svd.ref # big_SVD object computed from reference data.

# Make colours for the plot
x <- cbind(kgen_pcs.mx[,1:2],kgp_pops_colours)
eur <- x[x[,3]=="blue",]
non_eur <- x[x[,3]!="blue",]
x <- rbind(non_eur,eur)
y <- cbind(wecare_pcs.mx[, 1:2],"red")
z <- rbind(x,y)

# Plot 2 top PC-s
plot(z[,1:2],col=z[,3], xlab="PC1", ylab="PC2",
     main="Projecting WECARE Ampliseq to 1KGP\nusing only 161 variants")
text(5,0,"AFR")
text(-1,5,"EAS")
text(-1,1,"AMR/SAS")
text(-5,-6,"EUR")

legend("bottomright",legend="WECARE Ampliseq", pch=1, col ="red", bty="n")

# Clean-up
rm(x,y,z,eur,non_eur)

```

# End section

```{r}

ls()
Sys.time()
gc()

```
