---
title: "Ampliseq WECARE only: update the main dataset with relatedness and PCA results"
author: "AL"
date: "23 Dec 2021"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Summary

Add updated phenotype data frame that includes PC-s  

In case if there was PCA outliers and related samples, the script would also do the following (none of these was needed in this case):  

- Sync genotypes with the updated phenotypes:  
    + Removes related samples (if any) detected during PCA pre-processing  
    + Removes outliers detected in PCA (if any) from genotypes matrix  
- Remove non-polymorphic sites (if any) appeared after the samples filtering  
- Remove variants with call rate < 0.85 (the SKAT default threshold)  


Input data: 3,840 vars x 389 samples (194UBC + 195CBC)  
Output data: 3,840 vars x 389 samples (194UBC + 195CBC)  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_relatedness_and_pca")
scripts_folder <- file.path(base_folder,"scripts","s07_relatedness_and_pca")
setwd(scripts_folder)

```

# Load main dataset except phenotypes  

```{r}

# Load main data
load(file.path(base_folder,"data","s06_qc_filters","s04_check_samples.RData"))

# Check data
dim(phenotypes.df)
dim(genotypes.mx)
dim(variants.df)

# Update folders (just in case)
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_relatedness_and_pca")
scripts_folder <- file.path(base_folder,"scripts","s07_relatedness_and_pca")

# Remove phenotype data with related and outliers (if any) and without PC-s
colnames(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")
rm(phenotypes.df)

```

# Load updated phenotypes

```{r}

# Load updated phenotypes: without related and outliers (if any) and with PC-s
load(file.path(data_folder,"s06_phenotypes_with_PCs.RData")) 
dim(phenotypes_with_PCs.df)
colnames(phenotypes_with_PCs.df)
table(phenotypes_with_PCs.df$cc, useNA = "always")

# Update folders (just in case)
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_relatedness_and_pca")
scripts_folder <- file.path(base_folder,"scripts","s07_relatedness_and_pca")

```

# Sync genotypes with updated phenotypes

```{r}

# Check data
dim(phenotypes_with_PCs.df)
dim(genotypes.mx)

sum(phenotypes_with_PCs.df$sample.ID != colnames(genotypes.mx))

# Update genotypes (just in case?)
genotypes.mx <- genotypes.mx[,phenotypes_with_PCs.df$sample.ID]
dim(genotypes.mx)

# Rename phenotypes data frame
phenotypes.df <- phenotypes_with_PCs.df

# Clean-up
rm(phenotypes_with_PCs.df)

```

# Remove non-polymorphic sites

After samples filtering some variant sites may become non-polymorphic

```{r}

# Function to detect uniform numeric vector
uniformity_check.udf <- function(x){
  if (all(is.na(x))){"All_NA"}
  else if (min(x,na.rm=T)==max(x,na.rm=T)){"Uniform"}
  else {"Non-Uniform"}}
    
uniformity_check <- apply(genotypes.mx,1,uniformity_check.udf)
summary(as.factor(uniformity_check))
non_uniform_sites <- uniformity_check == "Non-Uniform"

# Update matrices and data-frame(s)
genotypes.mx <- genotypes.mx[non_uniform_sites,]
variants.df <- variants.df[non_uniform_sites,]

# Check result
dim(genotypes.mx)
dim(variants.df)

# Clean-up
rm(uniformity_check.udf,uniformity_check,non_uniform_sites)

```

# Variant call rates after filtering

Remove variants with call rate < 0.85 (the SKAT default threshold)

```{r}

# Function to calculate call rate
call_rate.udf <- function(x){1-sum(is.na(x))/length(x)}

# Call rates per variant
var_call_rate <- apply(genotypes.mx,1,call_rate.udf)
min(var_call_rate)

# Variants with low call rate
low_call_rate_variants <- var_call_rate < 0.85
sum(low_call_rate_variants)

# Update matrices and data-frame(s)
genotypes.mx <- genotypes.mx[!low_call_rate_variants,]
variants.df <- variants.df[!low_call_rate_variants,]

# Check result
dim(genotypes.mx)
dim(variants.df)

# Call rates per variant
var_call_rate <- apply(genotypes.mx,1,call_rate.udf)
min(var_call_rate)

hist(var_call_rate, main="Call rate per variant", 
     xlim=c(0,1), xlab="Call rates")
abline(v=0.85, lty=2, col="red")

plot(var_call_rate, ylim=c(0,1), main="Call rate per variant")
abline(h=0.85, lty=2, col="red")

# Clean-up
rm(var_call_rate, low_call_rate_variants)

```

# Sample call rates after filtering

```{r}

# Call rates per sample
case_call_rate <- apply(genotypes.mx,2,call_rate.udf)
min(case_call_rate)

hist(case_call_rate, main="Call rate per sample", 
     xlim=c(0,1),xlab="Call rates")
abline(v=0.5, lty=2, col="red")

plot(case_call_rate, ylim=c(0,1), main="Call rate per sample")
abline(h=0.5, lty=2, col="red")

low_call_rate_indices <- which(case_call_rate < 0.75)
low_call_rate_values <- case_call_rate[low_call_rate_indices]
low_call_rate_samples <- substr(names(low_call_rate_values),1,3)

low_call_rate_samples

text(x=low_call_rate_indices,y=low_call_rate_values,
     labels=low_call_rate_samples,pos=c(1,3,4),
     cex=0.7)

# Clean-up
rm(call_rate.udf,case_call_rate,
   low_call_rate_indices,low_call_rate_values,low_call_rate_samples)

```

# Genotypes missingness after filtering

```{r}

sum(is.na(genotypes.mx))/(nrow(genotypes.mx)*ncol(genotypes.mx))
genotypes.mx[1:5,1:5]

```

# Check data

```{r}

ls()

# Check objects dimensions
dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save(list=c("genotypes.mx","phenotypes.df","variants.df"),
     file=file.path(data_folder,"s08_update_main_dataset.RData"))

```

# End section

```{r}

ls()
Sys.time()
gc()

```
