---
title: "WES: BC11 proportion of carriers, WECARE only"
author: "Alexey Larionov"
date: "22 APr 2022"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Get ID-s of the carriers  
- Count carriers for cases and controls  
- Fisher and Prop tests for the carriers counts  
- Make a bar-plot for the carriers counts  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")

# Libraries
library(dplyr)

```

# Load data

```{r}

load(file.path(data_folder,"s03_allele_counts_bc11.RData"))

# Update foilders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

dim(bc11_variants.df)
dim(bc11_fiv.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

table(phenotypes.df$cc, useNA = "always")

# Check list of genes
bc11_genes

```

# Carriers ID-s and totals

```{r}

genotypes_bc11.mx <- genotypes.mx[bc11_fiv.df$ID,]
dim(genotypes_bc11.mx)

carriers_ids.udf <- function(x){paste(names(x)[which(x>0)], collapse=",")}
apply(genotypes_bc11.mx,1,carriers_ids.udf)

carriers_counts.udf <- function(x){sum(x>0, na.rm=T)}
carriers_counts <- apply(genotypes_bc11.mx,1,carriers_counts.udf)
carriers_counts

sum(carriers_counts)
sum(carriers_counts)/ncol(genotypes.mx)

rm(carriers_ids.udf, carriers_counts.udf, carriers_counts)

```

# Proportions of carriers

## Prepare indices

```{r}

table(phenotypes.df$cc)

controls <- phenotypes.df$cc == 0
cases <- phenotypes.df$cc == 1

bc11_fiv <- bc11_fiv.df$ID

```

## Controls

```{r}

gt_fiv_controls.mx <- genotypes.mx[bc11_fiv,controls]
dim(gt_fiv_controls.mx)

controls_carriers <- apply(gt_fiv_controls.mx,2,sum,na.rm=T) 
table(controls_carriers,useNA = "always")

controls_carriers_count <- sum(controls_carriers>0)
controls_total_count <- sum(controls) # sum(), NOT length() !
controls_carriers_fraction <- controls_carriers_count / controls_total_count

rm(gt_fiv_controls.mx,controls_carriers,controls)

```

## Cases

```{r}

gt_fiv_cases.mx <- genotypes.mx[bc11_fiv,cases]
dim(gt_fiv_cases.mx)

cases_carriers <- apply(gt_fiv_cases.mx,2,sum,na.rm=T) 
table(cases_carriers,useNA = "always")

cases_carriers_count <- sum(cases_carriers>0)
cases_total_count <- sum(cases) # sum(), NOT length !
cases_carriers_fraction <- cases_carriers_count / cases_total_count

rm(gt_fiv_cases.mx,cases_carriers,bc11_fiv,cases)

```

## Fisher test

For counts of carriers

```{r}

fisher_matrix <- matrix(c(cases_carriers_count, cases_total_count - cases_carriers_count,
                          controls_carriers_count, controls_total_count - controls_carriers_count),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("Cases","Controls")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(fisher_matrix)

```

## Prop test

For proportions of carriers

```{r}

pt <- prop.test(c(controls_carriers_count,cases_carriers_count),
          c(controls_total_count,cases_total_count))

pt

```

## Barplot

```{r}

# Prepare data for plot
proportions_of_carriers <- c(controls_carriers_fraction,cases_carriers_fraction)
names(proportions_of_carriers) <- c("UBC","CBC")
cc_counts <- table(phenotypes.df$cc)

# Barplot (keep mid-positions of bars)
mp <- barplot(proportions_of_carriers, ylim=c(0,max(proportions_of_carriers)+0.015), 
              main=paste("BC11 Proportion of PV carriers: WES WECARE-only\n",
                cc_counts[1],"controls vs ",cc_counts[2],
                " cases\nCrude counts Fisher p=",round(ft$p.value,3)),
              ylab="Proportion of PGV carriers")

# Add labels to plot
text(mp,proportions_of_carriers+0.007,round(proportions_of_carriers,3))

# Barplot with simplified title
mp <- barplot(proportions_of_carriers, ylim=c(0,max(proportions_of_carriers)+0.015), 
              main="Proportion of PGV carriers",
              ylab=NA)

# Add labels to plot
text(mp,proportions_of_carriers+0.007,round(proportions_of_carriers,3))

# Clean-up
rm(controls_carriers_count, controls_total_count, controls_carriers_fraction,
   cases_carriers_count, cases_total_count, cases_carriers_fraction,
   proportions_of_carriers, mp, cc_counts,
   pt,ft)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc11_variants.df)
dim(bc11_fiv.df)

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

table(phenotypes.df$cc, useNA = "always")

```

# Save results

```{r}

save.image(file.path(data_folder,"s04_proportin_of_carriers_bc11.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
