---
title: "Compare phenotypes of 2018, 2019 and 2021"
author: "AL"
date: "20 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

- Initially, Ampliseq phenotype data were provided by Xiaolin in e-mail to Marc of **29Aug2018**  
  file name: **Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  
  
- Then, the updated data were provided to me by Marc in e-mail of **07Jan2019**  
  file name: **2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  
  
- Finally an update was sent in **03Nov2021** to exclude 3 more samples and to swop numbers in one pair of samples  
  file name: **Correction0924201_2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**

This script reads the sets and compares them before use.  

Because of a problem with XLConnect:  
https://stackoverflow.com/questions/tagged/xlconnect?sort=active  
the worksheets containing phenotype data were saved as tab-separated text files to import to R.  

The comparison shows that 

1) Data of 2019 is identical to data of 2018, except for addition of 3 new columns:  
- family_history  
- setno  
- xrtbreast  

2) Data of 2021 is identical to data of 2019 except for:
- order of rows  
- removal of samples 391, 393 and 395  
- swap of samples 240 and 241  

All differences are expected (they were described in correspondence with Xiaolin)  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

#options(warnPartialMatchArgs = T, 
#        warnPartialMatchAttr = T, 
#        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s03_read_data")
scripts_folder <- file.path(base_folder,"scripts","s03_read_data")
setwd(scripts_folder)

```

# Read source data

## 2018

```{r}

data_file_2018 <- file.path(base_folder,"data","s01_phenotypes","txt","Phenotype data for exome seq_shipping manifest_sent_10 05 17.txt")

phenotypes_2018.df <- read.table(data_file_2018,
                                 header=T,sep="\t",quote="")
dim(phenotypes_2018.df)
colnames(phenotypes_2018.df)

rm(data_file_2018)

```

## 2019

```{r}

data_file_2019 <- file.path(base_folder,"data","s01_phenotypes","txt","2Phenotype data for exome seq_shipping manifest_sent_10 05 17.txt")

phenotypes_2019.df <- read.table(data_file_2019,
                                 header=T,sep="\t",quote="")
dim(phenotypes_2019.df)
colnames(phenotypes_2019.df)

rm(data_file_2019)

```

## 2021

```{r}

data_file_2021 <- file.path(base_folder,"data","s01_phenotypes","txt","Correction0924201_2Phenotype data for exome seq_shipping manifest_sent_10 05 17.txt")

phenotypes_2021.df <- read.table(data_file_2021,
                                 header=T,sep="\t",quote="")
dim(phenotypes_2021.df)
colnames(phenotypes_2021.df)

rm(data_file_2021)

```

# Compare 2018 and 2019

```{r}

all(phenotypes_2018.df == phenotypes_2019.df[,1:18])

```

# Compare 2019 and 2021

```{r}

identical(colnames(phenotypes_2019.df),colnames(phenotypes_2021.df))

phenotypes_2021.df <- phenotypes_2021.df[order(phenotypes_2021.df$Sample_num),]
rownames(phenotypes_2021.df) <- phenotypes_2021.df$Sample_num

phenotypes_2019.df <- phenotypes_2019.df[
  !phenotypes_2019.df$Sample_num %in% c(391,393,395),]

rownames(phenotypes_2019.df) <- phenotypes_2019.df$Sample_num

old240 <- phenotypes_2019.df[240,]
old241 <- phenotypes_2019.df[241,]
  
phenotypes_2019.df[240,] <- old241
phenotypes_2019.df[240,"Sample_num"] <- 240

phenotypes_2019.df[241,] <- old240 
phenotypes_2019.df[241,"Sample_num"] <- 241

identical(colnames(phenotypes_2019.df),colnames(phenotypes_2021.df))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
