---
title: "Add flag for Danish ampliseq samples"
author: "AL"
date: "20 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

- The final phenotype data were provided to me by Marc in e-mail of **03Nov2021**  
    + file name: **Correction0924201_2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  
- The information about Danish samples was sent by Xiaolin to Marc on **11Sep2019**  
    + file name: **Danish_samples_in_WES.xlsx**

In e-mail of **27Apr2020** Marc clarified that clinically actionable genes must not be analyzed in Danish cohort.  
So, these samples should be excluded from the current analysis because it aggregates variants in the known BC genes.  

This script flags Danish samples in ampliseq phenotype data.  

Because of a problem with XLConnect:  
https://stackoverflow.com/questions/tagged/xlconnect?sort=active  
the worksheets containing phenotype data were saved as tab-separated text files to import to R.  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

#options(warnPartialMatchArgs = T, 
#        warnPartialMatchAttr = T, 
#        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s03_read_data")
scripts_folder <- file.path(base_folder,"scripts","s03_read_data")
setwd(scripts_folder)

# Libraries
library(dplyr)

```

# Read source data

## 2021  

```{r}

# Read
data_file_2021 <- file.path(base_folder,"data","s01_phenotypes","txt","Correction0924201_2Phenotype data for exome seq_shipping manifest_sent_10 05 17.txt")
phenotypes_2021.df <- read.table(data_file_2021,
                                 header=T,sep="\t",quote = "")

# Explore
dim(phenotypes_2021.df)
str(phenotypes_2021.df)
phenotypes_2021.df[1:5,1:5]

# Order and add rownames
phenotypes_2021.df <- phenotypes_2021.df[
  order(phenotypes_2021.df$Sample_num),]

rownames(phenotypes_2021.df) <- phenotypes_2021.df$Sample_num

# Clean-up
rm(data_file_2021)

```

## Danish

```{r}

# Read
data_file <- file.path(base_folder,"data","s01_phenotypes","txt","Danish_samples_in_WES_round2.txt")
danish_samples.df <- read.table(data_file,
                                header=T,sep="\t",quote = "")

# Explore
dim(danish_samples.df)
str(danish_samples.df)
danish_samples.df[c(1,38,74),]

# Clean-up
rm(data_file)

```

# Add danish information

```{r}

danish <- phenotypes_2021.df$Sample_num %in% danish_samples.df$Sample_number
sum(danish)

phenotypes_2019_with_danish.df <- data.frame(phenotypes_2021.df,danish)
str(phenotypes_2019_with_danish.df)
sum(phenotypes_2019_with_danish.df$danish)

rm(danish_samples.df, phenotypes_2021.df, danish)

```

# Save result

```{r}

ls()
save.image(file.path(data_folder,"s03_flag_danish_samples.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
