---
title: "Ampliseq wecare NFFE: filter by variants call rate"
author: "Alexey Larionov"
date: "29 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

0.85 variant call rate is required for **each** NFE and WECARE for comparability of the data  

Remove **~69%** variants with low call rates (~ 31% of variants in wecare and 56% of variants in nffe).  

Call rate cut-off of 0.85 was selected for consistency with SKAT, which requires at least 85% call rate per variant  

Check call rates (per variant and per sample) before and after filtering.  

0.85 variant call rate is required for **each** NFE and WECARE for comparability of the data  

Input data: 6,157 variants x 587 samples (198 NFFE, 194 UBC & 195 CBC)  
Output data: 1,922 variants x 587 samples (198 NFFE, 194 UBC & 195 CBC)  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s02_ampliseq_wecare_nfe"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")
setwd(scripts_folder)

# Libraries
library(reshape2)
library(ggplot2)

# Call rate thresholds 
min_var_call_rate <- 0.85 # for filtering
min_sample_call_rate <- 0.5 # for estimate of the future filter effect

```

# Load data

```{r}

load(file.path(data_folder,"s02_filter_genotypes_by_alt_fraction.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s02_ampliseq_wecare_nfe"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc,useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Call rates before filtering

## per genotypes

```{r}

# Fraction of NA genotypes
sum(is.na(gt_add.mx))/(nrow(gt_add.mx)*ncol(gt_add.mx))
gt_add.mx[1:5,1:5]

```

## per variant

```{r}

# Function to calculate call rate
call_rate.udf <- function(x){1-sum(is.na(x))/length(x)}

# Make separate gq matrices for nfe and wecare
dim(gt_add.mx)
colnames(gt_add.mx)[c(389,390)]
genotypes_nfe.mx <- gt_add.mx[,390:587]
genotypes_wecare.mx <- gt_add.mx[,1:389]

dim(genotypes_nfe.mx)
dim(genotypes_wecare.mx)

genotypes_nfe.mx[1:5,1:5]
genotypes_wecare.mx[1:5,1:5]

# Call rates per variant (separately for NFE and WECARE !)
var_call_rates_nfe <- apply(genotypes_nfe.mx,1,call_rate.udf)
var_call_rates_wecare <- apply(genotypes_wecare.mx,1,call_rate.udf)

quantile(var_call_rates_nfe)
quantile(var_call_rates_wecare)

mean(var_call_rates_nfe)
mean(var_call_rates_wecare)

# Prepare data for ggplot hist
data.df <- data.frame(nfe=var_call_rates_nfe, wecare=var_call_rates_wecare)
molten.df <- melt(data.df)
colnames(molten.df) <- c("group","call_rate")
head(molten.df)
tail(molten.df)

# Make plot
ggplot(molten.df, aes(x=call_rate, fill=group)) + 
  geom_histogram(alpha=0.5, bins=15, position="identity", aes(y=..density..)) +
  geom_vline(xintercept=min_var_call_rate, linetype="dashed", color = "red") +
  labs(title="WECARE and NFFE",
       subtitle="before variant call rate filtering",
       x = "Variant call rate", y ="Density")

```

## per sample

```{r}

# Calculate call rates
call_rates <- apply(gt_add.mx,2,call_rate.udf)

# Plot
plot(call_rates, main="WECARE and NFFE\nbefore varaints call rate filtering",
     ylab="Call rate per sample", xlab="Sample No")
abline(h=min_sample_call_rate, lty=2, col="red")
abline(v=389.5, lty=2, col="red")

low_call_rate_indices <- which(call_rates < min_sample_call_rate)
low_call_rate_values <- call_rates[low_call_rate_indices]
low_call_rate_samples <- names(low_call_rate_values)

text(x=low_call_rate_indices,y=low_call_rate_values,
     labels=low_call_rate_samples,pos=4,cex=0.75)

# Clean-up
rm(call_rates,low_call_rate_indices,
   low_call_rate_values,low_call_rate_samples)

```

# Remove variants with low call rates

0.85 variant call rate is required for **each** NFE and WECARE for comparability of the data  

## Estimate proportion of varaints with low call rate (<85%)

```{r}

sum(var_call_rates_nfe < min_var_call_rate)/length(var_call_rates_nfe)
sum(var_call_rates_wecare < min_var_call_rate)/length(var_call_rates_wecare)

low_var_call_rates <- var_call_rates_nfe < min_var_call_rate | 
                      var_call_rates_wecare < min_var_call_rate
sum(low_var_call_rates)/nrow(gt_add.mx)

sum(low_var_call_rates)
sum(!low_var_call_rates)

# Clean-up
rm(var_call_rates_nfe, var_call_rates_wecare, data.df, molten.df,
   genotypes_nfe.mx, genotypes_wecare.mx)

```

## Remove variants

```{r}

gt_add.mx <- gt_add.mx[!low_var_call_rates,]
gt_num.mx <- gt_num.mx[!low_var_call_rates,]
gt_chr.mx <- gt_chr.mx[!low_var_call_rates,]
dp.mx <- dp.mx[!low_var_call_rates,]
variants.df <- variants.df[!low_var_call_rates,]

# Check result
dim(gt_add.mx)
dim(variants.df)

# Clean-up
rm(low_var_call_rates)

```

# Call rates after filtering

## per genotypes

```{r}

# Fraction of NA genotypes
sum(is.na(gt_add.mx))/(nrow(gt_add.mx)*ncol(gt_add.mx))
gt_add.mx[1:5,1:5]

```

## per variant

0.85 variant call rate is required for **each** NFE and WECARE for comparability of the data  

```{r}

# Make separate gq matrices for nfe and wecare
genotypes_nfe.mx <- gt_add.mx[,390:587]
genotypes_wecare.mx <- gt_add.mx[,1:389]

dim(genotypes_nfe.mx)
dim(genotypes_wecare.mx)

genotypes_nfe.mx[1:5,1:5]
genotypes_wecare.mx[1:5,1:5]

# Call rates per variant (separately for NFE and WECARE !)
var_call_rates_nfe <- apply(genotypes_nfe.mx,1,call_rate.udf)
var_call_rates_wecare <- apply(genotypes_wecare.mx,1,call_rate.udf)

quantile(var_call_rates_nfe)
quantile(var_call_rates_wecare)

mean(var_call_rates_nfe)
mean(var_call_rates_wecare)

# Prepare data for ggplot hist
data.df <- data.frame(nfe=var_call_rates_nfe, wecare=var_call_rates_wecare)
molten.df <- melt(data.df)
colnames(molten.df) <- c("group","call_rate")
head(molten.df)
tail(molten.df)

# Make plot
ggplot(molten.df, aes(x=call_rate, fill=group)) + 
  geom_histogram(alpha=0.5, bins=15, position="identity", aes(y=..density..)) +
  geom_vline(xintercept=min_var_call_rate, linetype="dashed", color = "red") +
  labs(title="WECARE and NFFE",
       subtitle="after variant call rate filtering",
       x = "Variant call rate", y ="Density")

# Look at minimal variant call rates
min(var_call_rates_nfe,var_call_rates_nfe)
var_call_rates_all <- apply(gt_add.mx,1,call_rate.udf)
min(var_call_rates_all)

# Clean-up
rm(var_call_rates_nfe, var_call_rates_wecare, data.df, molten.df,
   genotypes_nfe.mx, genotypes_wecare.mx, var_call_rates_all,
   min_var_call_rate)

```

## per sample

```{r}

# Calculate call rates
call_rates <- apply(gt_add.mx,2,call_rate.udf)

# Plot
plot(call_rates, main="WECARE and NFE\nafter variant call rate filtering",
     ylab="Call rate per sample", xlab="Sample No")
abline(h=min_sample_call_rate, lty=2, col="red")
abline(v=389.5, lty=2, col="red")

low_call_rate_indices <- which(call_rates < min_sample_call_rate)
low_call_rate_values <- call_rates[low_call_rate_indices]
low_call_rate_samples <- names(low_call_rate_values)

text(x=low_call_rate_indices,y=low_call_rate_values,
     labels=low_call_rate_samples,pos=4,cex=0.75)

# Clean-up
rm(call_rates,low_call_rate_indices,call_rate.udf,min_sample_call_rate,
   low_call_rate_values,low_call_rate_samples)

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)
dim(variants.df)
dim(phenotypes.df)

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Save results

```{r}

save.image(file.path(data_folder,"s03_filter_variants_by_call_rate.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
