---
title: "Ampliseq wecare NFFE: filter samples by call rate"
author: "Alexey Larionov"
date: "29 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary 

- Remove 1 NFFE sample because of the low call rate (<0.5)  
- Check for non-polymorphic sites after sample filtering  
- Explore mean dp, call rate and Ts/Tv per case before andf after sample filtering  
- Rename gt_add.mx to genotypes.mx  
- Remove gt_num, gt_char and dp matrices  

Mosr samples have Ts/Tv ratio > 2.  However, some samples had TS/TV 5; and some even >5.  No samples were excluded because of the high Ts/Tv ratio because of possible distortions introduced by ampliseq and relatively low number of variants  

Input data: 1,922 variants x 587 samples (198 NFFE, 194 UBC & 195 CBC)  
Output data: 1,922 variants x 586 samples (197 NFFE, 194 UBC & 195 CBC)  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s02_ampliseq_wecare_nfe"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")
setwd(scripts_folder)

# Libraries
library(stringr) # for str_split (in Ti/Tv check)

# Functions
source(file.path(scripts_folder, "f01_TsTv_ratio.R"))

# Set minimal call rate per sample
min_sample_call_rate <- 0.5

```

# Load data

```{r}

load(file.path(data_folder, "s03_filter_variants_by_call_rate.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s02_ampliseq_wecare_nfe"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")

```

## Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Explore samples before filtering

## dp per sample

```{r}

# Calculate mean depth per sample (sor simplicity, dont exclude missed genotypes)
mean_dp <- apply(dp.mx, 2, mean, na.rm=T)

# Plot mean depth
colnames(dp.mx[,c(389,390)])
plot(mean_dp, main="Mean depth", xlab="Samples", xaxt='n')
abline(v=389.5, lty=2, col="red")
text(c(175,500),200,c("Wecare","NFFE"))

quantile(mean_dp)
mean(mean_dp)

mean_dp_wecare <- apply(dp.mx[,phenotypes.df$cc!=-1], 2, mean, na.rm=T)
quantile(mean_dp_wecare)
mean(mean_dp_wecare)

mean_dp_nffe <- apply(dp.mx[,phenotypes.df$cc==-1], 2, mean, na.rm=T)
quantile(mean_dp_nffe)
mean(mean_dp_nffe)

# Clean-up 
rm(mean_dp,mean_dp_wecare,mean_dp_nffe)

```

## call_rates per sample

```{r}

# Function to get call rate
call_rate.udf <- function(x){sum(!is.na(x))/length(x)}

# Calculate mean depth
call_rates <- apply(gt_add.mx, 2, call_rate.udf)

# Plot mean depth
plot(call_rates, main="Call rates", xlab="Samples", xaxt='n')
abline(v=389.5, lty=2, col="red")
text(c(175,500), 0.4 ,c("Wecare","NFFE"))
abline(h=min_sample_call_rate, lty=2, col="red")

# Low call rate sample(s)
low_call_rate <- call_rates < min_sample_call_rate
sum(low_call_rate)
low_call_rate_samples <- call_rates[ low_call_rate ]

x <- which(low_call_rate)
y <- low_call_rate_samples
my_labs <- names(low_call_rate_samples)

text(x, y, labels=my_labs, cex=0.5, pos=3)

# Compare means/quantiles
quantile(call_rates)
mean(call_rates)

call_rates_wecare <- apply(gt_add.mx[,phenotypes.df$cc!=-1], 2, call_rate.udf)
quantile(call_rates_wecare)
mean(call_rates_wecare)

call_rates_nffe <- apply(gt_add.mx[,phenotypes.df$cc==-1], 2, call_rate.udf)
quantile(call_rates_nffe)
mean(call_rates_nffe)

# Clean-up (keep low_samples for later use in filtering)
rm(call_rate.udf, call_rates_wecare, call_rates_nffe,
   low_call_rate, low_call_rate_samples,x,y,my_labs)

```

## TsTv per sample

Needs stringr and TsTv_ratio udf 

```{r}

TsTv.mx <- matrix(ncol=4,nrow=0)
colnames(TsTv.mx) <- c("case","TsTv","Ts","Tv")

for(sample in colnames(gt_chr.mx)){

  #sample <- "147_S493_L008" # A standard sample
  #sample <- "108_S482_L008" # All genotypes are NA: 
  #   will be excluded later for loq call rate
  
  # Initialise result
  z <- c(NA,NA,NA)
  
  # Get vector of genotypes
  g <- gt_chr.mx[,sample]
  
  # If there are at least 100 genotypes
  if(sum(!is.na(g)) > 100){
    
    # Parse genotypes
    x <- str_split(g[!is.na(g)], "/")
    y <- t(matrix(unlist(x),nrow=2))
    
    # Calculate TsTv
    Ref <- y[,1]
    Alt <- y[,2]
    z <- TsTv_ratio(Ref,Alt)
    
  }

  # Write result
  TsTv.mx <- rbind(TsTv.mx,c(sample,z))

}

# Check result
dim(TsTv.mx)
TsTv.mx[1:10,]

# Convert to data frame
TsTv.df <- as.data.frame(TsTv.mx)
TsTv.df$TsTv <- as.numeric(TsTv.mx[,2])
TsTv.df$Ts <- as.numeric(TsTv.mx[,3])
TsTv.df$Tv <- as.numeric(TsTv.mx[,4])
str(TsTv.df)

quantile(TsTv.df$TsTv, na.rm=T)

# Check cases with NA ratio
sum(is.na(TsTv.df$TsTv))
TsTv.mx[is.na(TsTv.df$TsTv),]

# Make TsTv plot
plot(TsTv.df$TsTv, main="TsTv ratios per sample", 
     xlab="samples", xaxt='n', ylab="Ts/Tv")

abline(h=2, col="red", lty=2)
abline(h=5, col="red", lty=2)

abline(v=389.5, lty=2, col="red")
text(c(175,500), 6 ,c("Wecare","NFFE"))

# High TsTv samples
high_score <- TsTv.df$TsTv>5.5 & !is.na(TsTv.df$TsTv)
sum(high_score)
TsTv.df[high_score,]

# Low TsTv samples
low_score <- TsTv.df$TsTv<2 & !is.na(TsTv.df$TsTv)
sum(low_score)
TsTv.df[low_score,]

# Examples of normal TsTv samples
TsTv.df[200:205, ]

# Clean-up
rm(sample, g, x, y, z, Ref, Alt, TsTv.mx, TsTv.df,
   high_score, low_score)

```

# Filter samples by call rate

```{r}

# Filter
selected_samples <- call_rates >= min_sample_call_rate
sum(selected_samples)

sum(!selected_samples)
phenotypes.df$long_ids[!selected_samples]

gt_add.mx <- gt_add.mx[,selected_samples]
gt_num.mx <- gt_num.mx[,selected_samples]
gt_chr.mx <- gt_chr.mx[,selected_samples]
dp.mx <- dp.mx[,selected_samples]

phenotypes.df <- phenotypes.df[selected_samples,] 

# Check result
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)

dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Clean-up
rm(selected_samples,call_rates,min_sample_call_rate)

```

# Check non-polymorphic sites

After samples filtering some variant sites may become non-polymorphic

```{r}

# Function to detect uniform numeric vector
uniformity_check.udf <- function(x){
  if (all(is.na(x))){"All_NA"}
  else if (min(x,na.rm=T)==max(x,na.rm=T)){"Uniform"}
  else {"Non-Uniform"}}
    
uniformity_check <- apply(gt_add.mx,1,uniformity_check.udf)
summary(as.factor(uniformity_check))

# Clean-up
rm(uniformity_check.udf,uniformity_check)

```

# Explore samples after filtering

## dp per sample

```{r}

# Calculate mean depth per sample (sor simplicity, dont exclude missed genotypes)
mean_dp <- apply(dp.mx, 2, mean, na.rm=T)

# Plot mean depth
colnames(dp.mx[,c(389,390)])
plot(mean_dp, main="Mean depth", xlab="Samples", xaxt='n')
abline(v=389.5, lty=2, col="red")
text(c(175,500),200,c("Wecare","NFFE"))

# Compare means/quantiles
quantile(mean_dp)
mean(mean_dp)

mean_dp_wecare <- apply(dp.mx[,phenotypes.df$cc!=-1], 2, mean, na.rm=T)
quantile(mean_dp_wecare)
mean(mean_dp_wecare)

mean_dp_nffe <- apply(dp.mx[,phenotypes.df$cc==-1], 2, mean, na.rm=T)
quantile(mean_dp_nffe)
mean(mean_dp_nffe)

# Clean-up 
rm(dp.mx, mean_dp,mean_dp_wecare,mean_dp_nffe)

```

## call_rates per sample

```{r}

# Function to get call rate
call_rate.udf <- function(x){sum(!is.na(x))/length(x)}

# Calculate mean depth
call_rates <- apply(gt_add.mx, 2, call_rate.udf)

# Plot mean depth
plot(call_rates, main="Call rates", xlab="Samples", xaxt='n')
abline(v=389.5, lty=2, col="red")
text(c(175,500), 0.7 ,c("Wecare","NFFE"))

# Compare means/quantiles
quantile(call_rates)
mean(call_rates)

call_rates_wecare <- apply(gt_add.mx[,phenotypes.df$cc!=-1], 2, call_rate.udf)
quantile(call_rates_wecare)
mean(call_rates_wecare)

call_rates_nffe <- apply(gt_add.mx[,phenotypes.df$cc==-1], 2, call_rate.udf)
quantile(call_rates_nffe)
mean(call_rates_nffe)

# Clean-up (keep low_samples for later use in filtering)
rm(call_rate.udf, call_rates,call_rates_wecare,call_rates_nffe)

```

## TsTv per sample

Needs stringr and TsTv_ratio udf 

```{r}

TsTv.mx <- matrix(ncol=4,nrow=0)
colnames(TsTv.mx) <- c("case","TsTv","Ts","Tv")

for(sample in colnames(gt_chr.mx)){

  #sample <- "147_S493_L008" # A standard sample
  #sample <- "108_S482_L008" # All genotypes are NA: 
  #   will be excluded later for loq call rate
  
  # Initialise result
  z <- c(NA,NA,NA)
  
  # Get vector of genotypes
  g <- gt_chr.mx[,sample]
  
  # If there are at least 100 genotypes
  if(sum(!is.na(g)) > 100){
    
    # Parse genotypes
    x <- str_split(g[!is.na(g)], "/")
    y <- t(matrix(unlist(x),nrow=2))
    
    # Calculate TsTv
    Ref <- y[,1]
    Alt <- y[,2]
    z <- TsTv_ratio(Ref,Alt)
    
  }

  # Write result
  TsTv.mx <- rbind(TsTv.mx,c(sample,z))

}

# Check result
dim(TsTv.mx)
TsTv.mx[1:10,]

# Convert to data frame
TsTv.df <- as.data.frame(TsTv.mx)
TsTv.df$TsTv <- as.numeric(TsTv.mx[,2])
TsTv.df$Ts <- as.numeric(TsTv.mx[,3])
TsTv.df$Tv <- as.numeric(TsTv.mx[,4])
str(TsTv.df)

quantile(TsTv.df$TsTv, na.rm=T)

# Check cases with NA ratio
sum(is.na(TsTv.df$TsTv))
TsTv.mx[is.na(TsTv.df$TsTv),]

# Make TsTv plot
plot(TsTv.df$TsTv, main="TsTv ratios per sample", 
     xlab="samples", xaxt='n', ylab="Ts/Tv")

abline(h=2, col="red", lty=2)
abline(h=5, col="red", lty=2)

abline(v=389.5, lty=2, col="red")
text(c(175,500), 6 ,c("Wecare","NFFE"))

# High TsTv samples
high_score <- TsTv.df$TsTv>5.5 & !is.na(TsTv.df$TsTv)
sum(high_score)
TsTv.df[high_score,]

# Low TsTv samples
low_score <- TsTv.df$TsTv<2 & !is.na(TsTv.df$TsTv)
sum(low_score)
TsTv.df[low_score,]

# Examples of normal TsTv samples
TsTv.df[200:205, ]

# Clean-up
rm(sample, g, x, y, z, Ref, Alt, TsTv_ratio, TsTv.mx, TsTv.df,
   high_score, low_score, 
   gt_chr.mx)

```

# Rename and remove matrices

```{r}

genotypes.mx <- gt_add.mx
rm(gt_num.mx,gt_add.mx)

```

# Data summary

```{r}

# List objects
ls()

# Check sizes
dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency
sum(rownames(genotypes.mx) != rownames(variants.df))
sum(colnames(genotypes.mx) != rownames(phenotypes.df)) 

```

# Save data

```{r}

save.image(file.path(data_folder, "s04_filter_samples_by_call_rate.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
