---
title: "Ampliseq wecare only: filter variants by call rate"
author: "Alexey Larionov"
date: "23 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

Remove ~32% variants with low call rate.  

The 0.85 threshold is selected because SKAT requires at least 85% call rate per variant.  

Check call rates (per variant and per sample) before and after filtering.  

Input data: 5,624 variants x 389 samples (194UBC + 195CBC)  
Output data: 3,840 variants x 389 samples (194UBC + 195CBC)  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")
setwd(scripts_folder)

```

# Load data

```{r}

load(file.path(data_folder,"s02_filter_genotypes_by_alt_fraction.RData"))

# Update folders (just in case)
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s06_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s06_qc_filters")

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Genotypes before filtering

```{r}

# Fraction of NA genotypes
sum(is.na(gt_add.mx))/(nrow(gt_add.mx)*ncol(gt_add.mx))
gt_add.mx[1:5,1:5]

```

# Variant call rates before filtering

```{r}

# Function to calculate call rate
call_rate.udf <- function(x){1-sum(is.na(x))/length(x)}

# Call rates per variant
var_call_rate <- apply(gt_add.mx,1,call_rate.udf)

hist(var_call_rate, breaks=50, main="Call rate per variant", xlab="Call rates")
abline(v=0.85, lty=2, col="red")

plot(var_call_rate, main="Call rate per variant")
abline(h=0.85, lty=2, col="red")

```

# Sample call rates before filtering

```{r}

# Call rates per sample
case_call_rate <- apply(gt_add.mx,2,call_rate.udf)

hist(case_call_rate, xlim=c(0,1), main="Call rate per sample", xlab="Call rates")
abline(v=0.5, lty=2, col="red")

plot(case_call_rate, ylim=c(0,1), main="Call rate per sample")
abline(h=0.5, lty=2, col="red")

low_call_rate_indices <- which(case_call_rate < 0.5)
low_call_rate_values <- case_call_rate[low_call_rate_indices]
low_call_rate_samples <- substr(names(low_call_rate_values),1,3)

text(x=low_call_rate_indices,y=low_call_rate_values,
     labels=low_call_rate_samples,pos=c(1,3,1), cex=0.7)

low_call_rate_values

# Clean-up
rm(case_call_rate,
   low_call_rate_indices,low_call_rate_values,low_call_rate_samples)

```

# Filter by variants call rates

```{r}

# Check the fraction of variants to be removed
sum(var_call_rate < 0.85)/length(var_call_rate)

# Filter
selected_variants <- var_call_rate >= 0.85
sum(selected_variants)

gt_add.mx <- gt_add.mx[selected_variants,]
gt_num.mx <- gt_num.mx[selected_variants,]
gt_chr.mx <- gt_chr.mx[selected_variants,]
dp.mx <- dp.mx[selected_variants,]
variants.df <- variants.df[selected_variants,]

# Check result
dim(gt_add.mx)
dim(variants.df)

# Clean-up
rm(selected_variants)

```

# Genotypes after filtering

```{r}

# Fraction of NA genotypes
sum(is.na(gt_add.mx))/(nrow(gt_add.mx)*ncol(gt_add.mx))
gt_add.mx[1:5,1:5]

```

# Variant call rates after filtering

```{r}

# Call rates per variant
var_call_rate <- apply(gt_add.mx,1,call_rate.udf)

hist(var_call_rate, breaks=50, main="Call rate per variant", 
     xlim=c(0,1), xlab="Call rates")
abline(v=0.85, lty=2, col="red")

plot(var_call_rate, ylim=c(0,1),main="Call rate per variant")
abline(h=0.85, lty=2, col="red")

# Clean-up
rm(var_call_rate)

```

# Sample call rates after filtering

```{r}

# Call rates per sample
case_call_rate <- apply(gt_add.mx,2,call_rate.udf)

hist(case_call_rate, xlim=c(0,1), main="Call rate per sample", xlab="Call rates")
abline(v=0.5, lty=2, col="red")

plot(case_call_rate, ylim=c(0,1), main="Call rate per sample")
abline(h=0.5, lty=2, col="red")

low_call_rate_indices <- which(case_call_rate < 0.75)
low_call_rate_values <- case_call_rate[low_call_rate_indices]
low_call_rate_samples <- substr(names(low_call_rate_values),1,3)

text(x=low_call_rate_indices,y=low_call_rate_values,
     labels=low_call_rate_samples,pos=c(1,3,1), cex=0.7)

low_call_rate_values

# Clean-up
rm(call_rate.udf,case_call_rate,
   low_call_rate_indices,low_call_rate_values,low_call_rate_samples)

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(dp.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Save results

```{r}

save.image(file.path(data_folder,"s03_filter_variants_by_call_rate.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
