---
title: "Ampliseq: BC9 crude counts, WECARE NFFE"
author: "Alexey Larionov"
date: "30 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

Only 9 genes from the BC11 gene list were included in Ampliseq panel: BC8 + RAD51C.  

- Select functionally important variants in BC9  

Variants selection: 

**Known or predicted pathogenic**  

*Known pathogenic*  
- Annotated as pathogenic/likely pathogenic in ClinVar (v?)  

*Predicted pathogenic*  
- LoF (HIGH impact by VEP)  
- Consensus high grade missenses: deleterious by SIFT & probably_damaging by PolyPhen & CADD >=25  

**Rare**  
- Rare (AF in Wecare < 0.01)  

**Exclude known benign**  
- Excluding known benign in ClinVar  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_afterNov2021","ampliseq","s02_ampliseq_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")

# Libraries
library(dplyr)

```

# Load data

```{r}

load(file.path(data_folder,"s01_crude_counts_per_variant.RData"))

# Update working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_afterNov2021","ampliseq","s02_ampliseq_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
data_folder <- file.path(project_folder,"data","s08_crude_counts")

bc11_genes <- scan(file.path(data_folder,"bc11_genes.txt"),what="character")
# to do:  put bc11_genes.txt to scripts folder?

```

# Check data

```{r}

ls()

# Check objects dimensions
dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

# Check list of genes
bc11_genes

```

# Select variants 

## Keep only variants in the selected genes

Only 9 genes of the 11 were included into Ampliseq panel: 
ATM,CHEK2,TP53,NF1,NBN,CDH1,PTEN,STK11 & RAD51C.

Somehow, no variants in PTEN & STK11 were detected at all (i.e. no even common variants passed filters).  

```{r}

bc9_variants.df <- variants.df[variants.df$vep_SYMBOL %in% bc11_genes,]
dim(bc9_variants.df)

bc9_variants.df %>% 
  group_by(vep_SYMBOL) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))
# print(n=Inf)

```

## Select known (likely) pathogenic by ClinVar

```{r}

# Explore ClinVar annotations
bc9_variants.df %>% 
  group_by(CLNSIG) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

# Select known_pathogenic variants
known_pathogenic <- 
  grepl("Pathogenic",bc9_variants.df$CLNSIG) | 
  grepl("Likely_pathogenic",bc9_variants.df$CLNSIG)

sum(known_pathogenic)

```

## Predicted_lof-s

as defined by VEP IMPACT

```{r}

predicted_lof <- bc9_variants.df$vep_IMPACT == "HIGH"
sum(predicted_lof)

```

## Consensus high grade missenses  

### Explore SIFT and PolyPhen

```{r}

# SIFT
bc9_variants.df %>% 
  group_by(SIFT_call) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

# PolyPhen
bc9_variants.df %>% 
  group_by(PolyPhen_call) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))

```

### Explore CADD

```{r}

class(bc9_variants.df$vep_CADD_PHRED)
hist(bc9_variants.df$vep_CADD_PHRED, ylim=c(0,45),lab=T)
head(bc9_variants.df$vep_CADD_PHRED)
sum(is.na(bc9_variants.df$vep_CADD_PHRED))

# Check the variants missed CADD: both variants are of HIGH VEP-impact, 
# so they will be included anyway because of VEP-impact
bc9_variants.df[is.na(bc9_variants.df$vep_CADD_PHRED),
                c("vep_SYMBOL","vep_VARIANT_CLASS","vep_IMPACT","vep_Consequence",
                  "vep_Existing_variation","CLNSIG","vep_gnomAD_NFE_AF","vep_CADD_PHRED")]

```

### Consensus high grade missenses

```{r}

consensus_high_grade_missenses <- 
  bc9_variants.df$SIFT_call == "deleterious" & !is.na(bc9_variants.df$SIFT_call) & 
  bc9_variants.df$PolyPhen_call == "probably_damaging" & !is.na(bc9_variants.df$PolyPhen_call) & 
  bc9_variants.df$vep_CADD_PHRED >=25 & !is.na(bc9_variants.df$vep_CADD_PHRED)

sum(consensus_high_grade_missenses)

```

## Rare variants

```{r}

# Check missed AF data (just in case)
sum(is.na(bc9_variants.df$total_af))

# Rare in dataset
sum(bc9_variants.df$total_af <= 0.01, na.rm=T)
sum(bc9_variants.df$total_af >= 0.99, na.rm=T)

min(bc9_variants.df$total_af, na.rm=T)
max(bc9_variants.df$total_af, na.rm=T)

# Single benign variant with MAF < 0.01 and AF > 0.99 is irrelevant for the analysis
bc9_variants.df[bc9_variants.df$total_af>0.99,
                c("vep_SYMBOL","vep_VARIANT_CLASS","vep_IMPACT","vep_Consequence",
                  "vep_Existing_variation","CLNSIG","vep_gnomAD_NFE_AF","vep_CADD_PHRED")]

# Select rare variants
rare <- bc9_variants.df$total_af <= 0.01
sum(rare)

```

## Known benign

```{r}

bc9_variants.df %>% 
  group_by(CLNSIG) %>% 
  summarise(counts=n()) %>% 
  arrange(desc(counts))

known_benign <- 
  grepl("Benign",bc9_variants.df$CLNSIG) | 
  grepl("Likely_benign",bc9_variants.df$CLNSIG)

sum(known_benign)

```

## Combined functionaly important variants

```{r}

fiv <- (predicted_lof | known_pathogenic | consensus_high_grade_missenses) & 
        rare & 
        !(known_benign)

sum(fiv)

bc9_fiv.df <- bc9_variants.df[fiv,] %>% 
  select(ID,vep_SYMBOL,vep_Consequence,vep_IMPACT,CLNSIG,CLNREVSTAT,PolyPhen_call,SIFT_call,
         vep_MAX_AF,vep_gnomAD_NFE_AF,
         total_ac,total_an,total_af,
         nfe_ac,nfe_an,nfe_af,
         controls_ac,controls_an,controls_af,
         cases_ac,cases_an,cases_af)

#View(bc9_fiv.df)
bc9_fiv.df

write.table(bc9_fiv.df, 
            file.path(data_folder,"bc9_fiv_ampliseq_wecare_nffe.txt"),
            quote=F,sep ="\t",row.names=F,col.names=T)

# Clean-up
rm(predicted_lof,known_pathogenic,rare,fiv,consensus_high_grade_missenses,known_benign)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc9_variants.df)
dim(bc9_fiv.df)

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s02_select_bc9_pv.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
