---
title: "Ampliseq: BC9 proportion of carriers, WECARE NFFE"
author: "Alexey Larionov"
date: "30 Dec 2021"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Get ID-s of the carriers  
- Count carriers for nffe, cbc and ubc  
- Prop.trend test for the carriers counts  
- Make a bar-plot for the carriers counts  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_afterNov2021","ampliseq","s02_ampliseq_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")

# Libraries
library(dplyr)

```

# Load data

```{r}

load(file.path(data_folder,"s03_allele_counts_bc9.RData"))

# Update working folders (just in case)
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_afterNov2021","ampliseq","s02_ampliseq_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
data_folder <- file.path(project_folder,"data","s08_crude_counts")

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

dim(bc9_variants.df)
dim(bc9_fiv.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Carriers ID-s and totals

```{r}

genotypes_bc9.mx <- genotypes.mx[bc9_fiv.df$ID,]
dim(genotypes_bc9.mx)

carriers_ids.udf <- function(x){paste(names(x)[which(x>0)], collapse=",")}
apply(genotypes_bc9.mx,1,carriers_ids.udf)

carriers_counts.udf <- function(x){sum(x>0, na.rm=T)}
carriers_counts <- apply(genotypes_bc9.mx,1,carriers_counts.udf)
carriers_counts

sum(carriers_counts)
sum(carriers_counts)/ncol(genotypes.mx)

rm(carriers_ids.udf, carriers_counts.udf, carriers_counts, genotypes_bc9.mx)

```

# Proportions of carriers

## Prepare indices

```{r}

table(phenotypes.df$cc)

nffe <- phenotypes.df$cc == -1
ubc <- phenotypes.df$cc == 0
cbc <- phenotypes.df$cc == 1

bc9_fiv <- bc9_fiv.df$ID

```

## nffe

```{r}

gt_fiv_nffe.mx <- genotypes.mx[bc9_fiv,nffe]
dim(gt_fiv_nffe.mx)

nffe_carriers <- apply(gt_fiv_nffe.mx,2,sum,na.rm=T) 
table(nffe_carriers,useNA = "always")

nffe_carriers_count <- sum(nffe_carriers>0)
nffe_total_count <- sum(nffe) # sum, NOT length
nffe_carriers_fraction <- nffe_carriers_count / nffe_total_count
nffe_carriers_fraction

rm(gt_fiv_nffe.mx,nffe_carriers,nffe)

```

## ubc

```{r}

gt_fiv_ubc.mx <- genotypes.mx[bc9_fiv,ubc]
dim(gt_fiv_ubc.mx)

ubc_carriers <- apply(gt_fiv_ubc.mx,2,sum,na.rm=T) 
table(ubc_carriers,useNA = "always")

ubc_carriers_count <- sum(ubc_carriers>0)
ubc_total_count <- sum(ubc) # sum, NOT length
ubc_carriers_fraction <- ubc_carriers_count / ubc_total_count
ubc_carriers_fraction

rm(gt_fiv_ubc.mx,ubc_carriers,ubc)

```

## cbc

```{r}

gt_fiv_cbc.mx <- genotypes.mx[bc9_fiv,cbc]
dim(gt_fiv_cbc.mx)

cbc_carriers <- apply(gt_fiv_cbc.mx,2,sum,na.rm=T) 
table(cbc_carriers,useNA = "always")

cbc_carriers_count <- sum(cbc_carriers>0)
cbc_total_count <- sum(cbc) # sum, NOT length
cbc_carriers_fraction <- cbc_carriers_count / cbc_total_count
cbc_carriers_fraction

rm(gt_fiv_cbc.mx,cbc_carriers,bc9_fiv,cbc)

```

## Prop trend test

For proportions of carriers

```{r}

pt <- prop.trend.test(c(nffe_carriers_count,ubc_carriers_count,cbc_carriers_count),
          c(nffe_total_count,ubc_total_count,cbc_total_count))

pt

rm(nffe_carriers_count,ubc_carriers_count,cbc_carriers_count,
   nffe_total_count,ubc_total_count,cbc_total_count,pt)

```

## Barplot

```{r}

# Prepare data for plot
proportions_of_carriers <- c(nffe_carriers_fraction,ubc_carriers_fraction,cbc_carriers_fraction)
names(proportions_of_carriers) <- c("NFFE","UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(proportions_of_carriers, 
              ylim=c(0,max(proportions_of_carriers)+0.01), 
              main="Proportion of PV carriers",
              ylab="Proportion of PV carriers")

# Add labels to plot
text(mp,proportions_of_carriers+0.005,round(proportions_of_carriers,3))

# Clean-up
rm(nffe_carriers_fraction,ubc_carriers_fraction,cbc_carriers_fraction,
   proportions_of_carriers, mp)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc9_variants.df)
dim(bc9_fiv.df)

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s04_proportin_of_carriers_bc9.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
