---
title: "WES: BC11 OR by latency"
author: "Alexey Larionov"
date: "22 Apr 2022"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Summary

Calculate CBC/UBC OR for different age groups  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s11_latency_or")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s11_latency_or")
#dir.create(data_folder)

# Libraries
#library(dplyr)

```

# Load data

```{r}

load(file.path(project_folder,"data","s08_crude_counts","s03_allele_counts_bc11.RData"))

# Update folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s01_wes_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s11_latency_or")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s11_latency_or")

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

dim(bc11_variants.df)
dim(bc11_fiv.df)

```

# Indices

## UBC / CBC

```{r}

table(phenotypes.df$cc, useNA = "always")

ubc <- phenotypes.df$cc == 0
cbc <- phenotypes.df$cc == 1

sum(ubc)
sum(cbc)

```

## Latency

```{r}

x <- phenotypes.df$age_ref - phenotypes.df$age_dx
plot(x, phenotypes.df$rstime,
     main="RS time vs difference of Dx and Rs",
     xlab="Difference of Dx and Rs",
     ylab="RS time")

z <- which(x - phenotypes.df$rstime > 1)
s <- phenotypes.df$sample.ID[z]
text(x[z], phenotypes.df$rstime[z], s, cex=0.5, pos=4)

y <- x - phenotypes.df$rstime
plot(y)
text(z, y[z], s, cex=0.5, pos=4)

# Note: "P1_C05" & "P2_H08" are not B11 FIV carriers
bad_rstime <- phenotypes.df$sample.ID %in% c("P1_C05","P2_H08")

hist(phenotypes.df$rstime[!bad_rstime],
     main = "WES: RS time distribution")

abline(v=5, col="red", lty=2)

early <- phenotypes.df$rstime <= 5 & !bad_rstime
late <- phenotypes.df$rstime > 5 & !bad_rstime

sum(early)
sum(late)
sum(early, late)

rm(x,y,z,s,bad_rstime)

```

## BC11 FIVs

```{r}

bc11_fiv_ids <- bc11_fiv.df$ID
bc11_fiv_count <- nrow(bc11_fiv.df)

```

# All samples

Here the counts are calculated directly from genotypes.mx  

For all sampes, it could be extracted from bc11_fiv.df  

However, bc11_fiv.df could not be used later because  
it doesn't have age-specific counts for CBC and UBC  

## UBC ac & an

```{r}

ubc_ac <- sum(genotypes.mx[bc11_fiv_ids,ubc], na.rm=T)

ubc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids,ubc]))*2
ubc_mean_an <- ubc_total_an / bc11_fiv_count

ubc_mean_af <- ubc_ac / ubc_total_an
ubc_aggregated_af <- ubc_ac / ubc_mean_an

rm(ubc_mean_an)

```

## CBC ac & an

```{r}

cbc_ac <- sum(genotypes.mx[bc11_fiv_ids,cbc], na.rm=T)

cbc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids,cbc]))*2
cbc_mean_an <- cbc_total_an / bc11_fiv_count

cbc_mean_af <- cbc_ac / cbc_total_an
cbc_aggregated_af <- cbc_ac / cbc_mean_an

rm(cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(ubc_ac, ubc_total_an - ubc_ac,
                          cbc_ac, cbc_total_an - cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("UBC","CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(ubc_ac,ubc_total_an,
   cbc_ac,cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(ubc_mean_af,cbc_mean_af)
names(afs) <- c("UBC","CBC")

ubc_count <- sum(phenotypes.df$cc == 0)
cbc_count <- sum(phenotypes.df$cc == 1)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("BC11 Mean AF: WES all ages\n",
                ubc_count," UBC vs ",cbc_count," CBC\n",
                "Crude counts Fisher p=",round(ft$p.value,3)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(ubc_mean_af, cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(ubc_aggregated_af,cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("BC11 Aggregated AF: WES all ages\n",
                ubc_count," UBC vs ",cbc_count," CBC\n",
                "Crude counts Fisher p=",round(ft$p.value,3)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,3))

# Clean-up
rm(ubc_aggregated_af, cbc_aggregated_af, afs, mp, 
   ubc_count, cbc_count, ft)

```

# Early participants

## Early UBC ac & an

```{r}

early_ubc_ac <- sum(genotypes.mx[bc11_fiv_ids, early & ubc], na.rm=T)

early_ubc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids, early & ubc]))*2
early_ubc_mean_an <- early_ubc_total_an / bc11_fiv_count

early_ubc_mean_af <- early_ubc_ac / early_ubc_total_an
early_ubc_aggregated_af <- early_ubc_ac / early_ubc_mean_an

rm(early_ubc_mean_an)

```

## Early CBC ac & an

```{r}

early_cbc_ac <- sum(genotypes.mx[bc11_fiv_ids, early & cbc], na.rm=T)

early_cbc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids, early & cbc]))*2
early_cbc_mean_an <- early_cbc_total_an / bc11_fiv_count

early_cbc_mean_af <- early_cbc_ac / early_cbc_total_an
early_cbc_aggregated_af <- early_cbc_ac / early_cbc_mean_an

rm(early_cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(early_ubc_ac, early_ubc_total_an - early_ubc_ac,
                          early_cbc_ac, early_cbc_total_an - early_cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("early UBC","early CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(early_ubc_ac,early_ubc_total_an,
   early_cbc_ac,early_cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(early_ubc_mean_af,early_cbc_mean_af)
names(afs) <- c("UBC","CBC")

early_ubc_count <- sum(phenotypes.df$cc == 0 & early)
early_cbc_count <- sum(phenotypes.df$cc == 1 & early)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("BC11 Mean AF: Early WES (<=5)\n",
                early_ubc_count," UBC vs ",early_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(early_ubc_mean_af, early_cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(early_ubc_aggregated_af,early_cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("BC11 Aggregated AF: Early WES (<=5)\n",
                early_ubc_count," UBC vs ",early_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Barplot with simplified title
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("Latency of 5 years or less\n",
                early_ubc_count," UBC vs ",early_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,2),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Clean-up
rm(early_ubc_aggregated_af, early_cbc_aggregated_af, afs, mp, 
   early_ubc_count, early_cbc_count, ft)

```

# Late participants

## Late UBC ac & an

```{r}

late_ubc_ac <- sum(genotypes.mx[bc11_fiv_ids, late & ubc], na.rm=T)

late_ubc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids, late & ubc]))*2
late_ubc_mean_an <- late_ubc_total_an / bc11_fiv_count

late_ubc_mean_af <- late_ubc_ac / late_ubc_total_an
late_ubc_aggregated_af <- late_ubc_ac / late_ubc_mean_an

rm(late_ubc_mean_an)

```

## Late CBC ac & an

```{r}

late_cbc_ac <- sum(genotypes.mx[bc11_fiv_ids, late & cbc], na.rm=T)

late_cbc_total_an <- sum(!is.na(genotypes.mx[bc11_fiv_ids, late & cbc]))*2
late_cbc_mean_an <- late_cbc_total_an / bc11_fiv_count

late_cbc_mean_af <- late_cbc_ac / late_cbc_total_an
late_cbc_aggregated_af <- late_cbc_ac / late_cbc_mean_an

rm(late_cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(late_ubc_ac, late_ubc_total_an - late_ubc_ac,
                          late_cbc_ac, late_cbc_total_an - late_cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("Late UBC","Late CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(late_ubc_ac,late_ubc_total_an,
   late_cbc_ac,late_cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(late_ubc_mean_af,late_cbc_mean_af)
names(afs) <- c("UBC","CBC")

late_ubc_count <- sum(phenotypes.df$cc == 0 & late)
late_cbc_count <- sum(phenotypes.df$cc == 1 & late)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("BC11 Mean AF: Late WES (>5)\n",
                late_ubc_count," UBC vs ",late_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(late_ubc_mean_af, late_cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(late_ubc_aggregated_af,late_cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("BC11 Aggregated AF: Late WES (>5)\n",
                late_ubc_count," UBC vs ",late_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Barplot with simplified title
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("Latency more than 5 years\n",
                late_ubc_count," UBC vs ",late_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,2),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Clean-up
rm(late_ubc_aggregated_af, late_cbc_aggregated_af, afs, mp, 
   late_ubc_count, late_cbc_count, ft,
   cbc,ubc,late,early,bc11_fiv_count,bc11_fiv_ids)

```

# Data summary

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

table(phenotypes.df$cc, useNA = "always")

dim(bc11_variants.df)
dim(bc11_fiv.df)

```

# Save results

```{r}

save.image(file.path(data_folder,"s01_latency_or_bc11.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
