---
title: "WES WECARE-NFFE BC11: Impute Weight and Aggregate variants for POLR"
author: "Alexey Larionov"
date: "10 Mar 2022"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary  

Perform SKAT-style data pre-processing and aggregation  
Preprocessing steps done in the following order:  

- Read and check bc8 data  
- *No need in inverting genotypes* because there is no variants with AFs > 0.5  
- Impute missed genotypes (2 x total AF = mean genotype, as "fixed" imputation in SKAT)  
- Weight variants by total AFs (dbeta[AF,1,25], default weight in SKAT, *assumes AF < 0.5* !!!)  
- Aggregate preprocessed (imputed-weighted) variants in bc11 genes (just summation: as in "burden" SKAT style)  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s09_polr")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s09_polr")
#dir.create(data_folder)

# Load libraries
library(dplyr)

```

# Load data

```{r}

load(file.path(project_folder,"data","s08_crude_counts","s03_allele_counts_bc11.RData"))

# Update working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s09_polr")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s09_polr")

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc11_variants.df)
dim(bc11_fiv.df)

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Select bc11 genotypes

```{r}

bc11_genotypes.mx <- genotypes.mx[bc11_fiv.df$ID,]
dim(bc11_genotypes.mx)
sum(is.na(bc11_genotypes.mx))

```

# Impute missed data

Substitute NA with mean genotype per variant (= duplicated **total** AFs).  
Equivalent to the "fixed" imputation approach in SKAT (see impute.method parameter for SKAT function).  
Note: No AFs > 0.5 !!!

```{r}

# Check AFs
max(bc11_fiv.df$total_af)

# Prepare matrix for output
bc11_genotypes_imp.mx <- bc11_genotypes.mx

# Fraction of missed data 
sum(is.na(bc11_genotypes_imp.mx))
sum(is.na(bc11_genotypes_imp.mx))/
  (nrow(bc11_genotypes_imp.mx)*ncol(bc11_genotypes_imp.mx))

# Impute missed data
for(var in 1:nrow(bc11_genotypes_imp.mx)){

    # Get AF
    af <- bc11_fiv.df[var,"total_af"]
    
    # Impute missed genotypes
    genotypes <- bc11_genotypes_imp.mx[var,]
    2*af -> genotypes[is.na(genotypes)]
    genotypes -> bc11_genotypes_imp.mx[var,]

    # Clean-up
    rm(af, genotypes)
}

# Explore imputed genotypes
sum(is.na(bc11_genotypes_imp.mx))
hist(bc11_genotypes_imp.mx, lab=T, ylim=c(0,9e3))
hist(bc11_genotypes_imp.mx[bc11_genotypes_imp.mx>0], lab=T, ylim=c(0,210),
     main="Histogram of genotypes\n(show only imputed and alts)")
min(bc11_genotypes_imp.mx[bc11_genotypes_imp.mx>0])

# Clean-up
rm(var)

```

# Weight variants by af

Weight by **total** AF using dbeta(AF,1,25)  
Similar to the default weighting in SKAT  
(see parameter weights.beta=c[1,25] for SKAT function)  

```{r}

# Prepare tables for output
bc11_fiv.df <- data.frame(bc11_fiv.df, weight=NA)
bc11_genotypes_imp_wt.mx <- bc11_genotypes_imp.mx

# Apply weighting
for(var in 1:nrow(bc11_genotypes_imp_wt.mx)){
 
  # Calculate weight
  weight <- dbeta(bc11_fiv.df[var, "total_af"], 1, 25)
  
  # Write weight to variants table
  weight -> bc11_fiv.df[var,"weight"]
  
  # Apply weights to genotypes table
  weight * bc11_genotypes_imp.mx[var,] -> bc11_genotypes_imp_wt.mx[var,]
  
  # Clean-up
  rm(weight)

}

# Explore weights
sum(is.na(bc11_fiv.df$weight))
mean(bc11_fiv.df$weight)
hist(bc11_fiv.df$weight, lab=T, ylim=c(0,15))

# Explore weighted genotypes
min(bc11_genotypes_imp_wt.mx)
min(bc11_genotypes_imp_wt.mx[bc11_genotypes_imp_wt.mx>0])
max(bc11_genotypes_imp_wt.mx)

hist(bc11_genotypes_imp_wt.mx, lab=T, ylim=c(0,9e3))
hist(bc11_genotypes_imp_wt.mx[bc11_genotypes_imp_wt.mx>0], lab=T, ylim=c(0,210), main="Histogram of genotypes\n(showing only imputed and alts)")

hist(bc11_genotypes_imp_wt.mx[bc11_genotypes_imp_wt.mx>0.5], lab=T, ylim=c(0,15), main="Histogram of genotypes\n(showing only alts)")

# clean-up
rm(var)

```

# Aggregate imputed weighted genotypes

```{r}

bc11_genotypes_imp_wt_aggregated <- apply(bc11_genotypes_imp_wt.mx,2,sum)

bc11_genotypes_imp_wt_aggregated[1:5]

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)

table(phenotypes.df$cc, useNA = "always")

dim(bc11_genotypes.mx)
dim(bc11_genotypes_imp.mx)
dim(bc11_genotypes_imp_wt.mx)

length(bc11_genotypes_imp_wt_aggregated)

length(bc11_genes)

dim(bc11_fiv.df)
dim(bc11_variants.df)

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

sum(names(bc11_genotypes_imp_wt_aggregated) != phenotypes.df$sample.ID)

```

# Save data

```{r}

save.image(file.path(data_folder, "s01_impute_weight_aggregate_bc11.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
