---
title: "WES WECARE-NFFE BC11: MASS POLR"
author: "Alexey Larionov"
date: "10 Mar 2022"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Reads the data, which is already imputed, weighted and aggregated per bc8 gene (there was no need in inverting genotypes).  
- Performs trend test for NFE < UBC < CBC using POLR function from MASS library.  

MASS POLR is a trend test within regression framework using proportional odds regression model. Thus, it allows the covariates, which are **3** top eigenvectors in this case.  

POLR or multinomial-logistic regression were advised by David Conti (e-mail of 18Nov2016). Initially, I tested both.  Then I excluded multinomuial because it was messy and, in general, multinomial is less sensitive/suitable for ordered categories.  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s09_polr")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s09_polr")

# Libraries
library(MASS) # for proportional odds regression (polr)

```

# Load data

```{r}

load(file.path(data_folder, "s01_impute_weight_aggregate_bc11.RData"))

# Update folders
# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","final_analysis_2021","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s09_polr")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s09_polr")

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)

table(phenotypes.df$cc, useNA = "always")

dim(bc11_genotypes.mx)
dim(bc11_genotypes_imp.mx)
dim(bc11_genotypes_imp_wt.mx)

length(bc11_genotypes_imp_wt_aggregated)

length(bc11_genes)

dim(bc11_fiv.df)
dim(bc11_variants.df)

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

sum(names(bc11_genotypes_imp_wt_aggregated) != phenotypes.df$sample.ID)

```

# POLR

## Prepare data for the model

### Outcomes

Make an ordered (!) factor

```{r}

# Check data
table(phenotypes.df$cc)

# Shift [-1,0,1] to [1,2,3]
# because natural numbers may better fit to ordered factors later 
Y <- phenotypes.df$cc + 2 

# Convert to ordered factor
Y <- ordered(Y, levels=c(1,2,3), labels=c("nffe", "ubc", "cbc")) 

# Add names (just in case)
names(Y) <- as.vector(phenotypes.df$sample.ID) 

# Check result
str(Y)
summary(Y)
Y[c(1,300,500)]
sum(is.na(Y)) # 0
Y[1:5]

```

### Eigenvectors

```{r}

E <- as.matrix(phenotypes.df[,c("pc1","pc2","pc3")])  
E[1:5,]
sum(is.na(E)) # 0

```

### Genotypes

```{r}

G <- bc11_genotypes_imp_wt_aggregated
sum(is.na(G))
length(G)
G[1:5]

```

## POLR test

### Null model (E)

Includes only Eigenvectors

```{r}

polr_E <- polr(Y ~ E)

```

### Complete model (GE)

Includes Genotypes and Eigenvectors

```{r}

polr_GE <- polr(Y ~ G + E)

```

### Null (E) vs Complete (GE)

Compute polr p-value for Genotypes

```{r}

polr_anova <- anova(polr_E, polr_GE)
polr_anova

polr_p <- polr_anova["2","Pr(Chi)"]
polr_p

# Clean-up
rm(Y, E, G, polr_p)

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)

table(phenotypes.df$cc, useNA = "always")

dim(bc11_genotypes.mx)
dim(bc11_genotypes_imp.mx)
dim(bc11_genotypes_imp_wt.mx)

length(bc11_genotypes_imp_wt_aggregated)

length(bc11_genes)

dim(bc11_fiv.df)
dim(bc11_variants.df)

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

sum(names(bc11_genotypes_imp_wt_aggregated) != phenotypes.df$sample.ID)

```

# Save data

```{r}

save.image(file.path(data_folder, "s02_mass_polr_bc11.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
