---
title: "Ampliseq: exclude Danish"
author: "Alexey Larionov"
date: "21 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary  

Remove 125 danish samples. Then exclude 1,118 non-polymorphic sites appeared after the samples removal.    

Input data: 12,816 vars x 712 samples (257UBC + 257CBC + 198NFFE)  
Output data: 11,698 vars x 587 samples (194UBC + 195CBC + 198NFFE)  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clean-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s04_select_samples")
scripts_folder <- file.path(base_folder,"scripts","s04_select_samples")
setwd(scripts_folder)

```

# Read data  

```{r}

# Load data
load(file.path(data_folder,"s03_exclude_BRCA1_BCRA2_PALB2_carriers.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s04_select_samples")
scripts_folder <- file.path(base_folder,"scripts","s04_select_samples")

```

# Check data

```{r}

ls()

dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)

dim(ad.mx)
dim(alt.mx)
dim(ref.mx)

dim(dp.mx)
dim(gq.mx)

dim(meta.df)

dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

dim(variants.df)

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Remove danish samples  

```{r}

# Check data
table(phenotypes.df$cc, useNA = "always")
table(phenotypes.df$danish, useNA = "always")
table(phenotypes.df[,c("cc","danish")], useNA = "always")

# Remove danish
sum(phenotypes.df$danish)
retained_samples <- ! phenotypes.df$danish
sum(retained_samples)

gt_add.mx <- gt_add.mx[,retained_samples]
gt_num.mx <- gt_num.mx[,retained_samples]
gt_chr.mx <- gt_chr.mx[,retained_samples]

ad.mx <- ad.mx[,retained_samples]
alt.mx <- alt.mx[,retained_samples]
ref.mx <- ref.mx[,retained_samples]

dp.mx <- dp.mx[,retained_samples]
gq.mx <- gq.mx[,retained_samples]

phenotypes.df <- phenotypes.df[retained_samples,]

# Check result
sum(colnames(gt_add.mx) != rownames(phenotypes.df))
dim(gt_add.mx)
dim(phenotypes.df)

gt_add.mx[1:5,1:5]

phenotypes.df[1:5,1:5]
phenotypes.df[583:587,1:5]

table(phenotypes.df$cc, useNA = "always")
table(phenotypes.df$danish, useNA = "always")
table(phenotypes.df[,c("cc","danish")], useNA = "always")

# Clean-up
rm(retained_samples)

```

# Remove non-polymorphic sites

```{r}

# Function to detect uniform numeric vector
uniformity_check.udf <- function(x){
  if (all(is.na(x))){"All_NA"}
  else if (min(x,na.rm=T)==max(x,na.rm=T)){"Uniform"}
  else {"Non-Uniform"}}

# Make the filter
uniformity_check <- apply(gt_add.mx,1,uniformity_check.udf)
summary(as.factor(uniformity_check))
non_uniform_sites <- uniformity_check == "Non-Uniform"

# Remove non-polymorphic variants
gt_add.mx <- gt_add.mx[non_uniform_sites,]
gt_num.mx <- gt_num.mx[non_uniform_sites,]
gt_chr.mx <- gt_chr.mx[non_uniform_sites,]

ad.mx <- ad.mx[non_uniform_sites,]
alt.mx <- alt.mx[non_uniform_sites,]
ref.mx <- ref.mx[non_uniform_sites,]

dp.mx <- dp.mx[non_uniform_sites,]
gq.mx <- gq.mx[non_uniform_sites,]

variants.df <- variants.df[non_uniform_sites,]

# Check result
dim(gt_add.mx)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")
table(phenotypes.df$danish, useNA = "always")
table(phenotypes.df[,c("cc","danish")], useNA = "always")

# Clean-up
rm(uniformity_check.udf, uniformity_check, non_uniform_sites)

```

# Check data

```{r}

ls()

dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)

dim(ad.mx)
dim(alt.mx)
dim(ref.mx)

dim(dp.mx)
dim(gq.mx)

dim(meta.df)

dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

dim(variants.df)

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Save results

```{r}

save.image(file.path(data_folder, "s04_exclude_danish.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
