---
title: "WES: BC11 crude allele counts, WECARE-NFFE"
author: "Alexey Larionov"
date: "22 Apr 2022"
output:
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Calculate aggregated crude counts: totals, and separately for nffe, cases and controls  
- Trend test for the crude counts  

**Aggregated** AF is defined as **total** AC over **mean** AN in the studied variants  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Clean up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

# Folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")
# folder already exists and contains a text file with BC11 genes

# Libraries
library(dplyr)

```

# Load data

```{r}

load(file.path(data_folder,"s02_select_bc11_pv.RData"))

# Update working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","wes","s02_wes_wecare_nfe")
scripts_folder <- file.path(project_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s08_crude_counts")
# folder already exists and contains a text file with BC11 genes

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc11_variants.df)
dim(bc11_fiv.df)

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

bc11_genes

```

# Aggregate crude counts

## Total

```{r}

total_ac <- sum(bc11_fiv.df$total_ac) 

total_an <- sum(bc11_fiv.df$total_an)
mean_an <- total_an / nrow(bc11_fiv.df)

mean_af <- total_ac / total_an
aggregated_af <- total_ac / mean_an

mean_af
aggregated_af

rm(total_ac,total_an,mean_an,mean_af,aggregated_af)

```

## NFFE

```{r}

nfe_ac <- sum(bc11_fiv.df$nfe_ac) 

nfe_total_an <- sum(bc11_fiv.df$nfe_an)
nfe_mean_an <- nfe_total_an / nrow(bc11_fiv.df)

nfe_mean_af <- nfe_ac / nfe_total_an
nfe_aggregated_af <- nfe_ac / nfe_mean_an

rm(nfe_mean_an)

```

## UBC

```{r}

controls_ac <- sum(bc11_fiv.df$controls_ac) 

controls_total_an <- sum(bc11_fiv.df$controls_an)
controls_mean_an <- controls_total_an / nrow(bc11_fiv.df)

controls_mean_af <- controls_ac / controls_total_an
controls_aggregated_af <- controls_ac / controls_mean_an

rm(controls_mean_an)

```

## CBC

```{r}

cases_ac <- sum(bc11_fiv.df$cases_ac)

cases_total_an <- sum(bc11_fiv.df$cases_an)
cases_mean_an <- cases_total_an / nrow(bc11_fiv.df)

cases_mean_af <- cases_ac / cases_total_an
cases_aggregated_af <- cases_ac / cases_mean_an

rm(cases_mean_an)

```

## Proportions trend test

nfe-ubc-cbc

```{r}

ac_prop_trend <- c(nfe_ac,controls_ac,cases_ac)
an_prop_trend <- c(nfe_total_an,controls_total_an,cases_total_an)
  
trend_test <- prop.trend.test(ac_prop_trend, an_prop_trend)
trend_test

rm(nfe_ac,controls_ac,cases_ac,
   nfe_total_an,controls_total_an,cases_total_an,
   ac_prop_trend, an_prop_trend)

```

## Barplot

### Mean AF

```{r}

afs <- c(nfe_mean_af,controls_mean_af,cases_mean_af)
names(afs) <- c("NFFE","UBC","CBC")

cc_counts <- table(phenotypes.df$cc)
cc_counts

mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("BC11: WES WECARE-NFFE\n",
                         cc_counts[1],"1KGP-NFFE,",
                         cc_counts[2],"Controls and",
                         cc_counts[3],"Cases\nCrude counts trend p=",round(trend_test$p.value,3)), 
              ylab="Mean AF in selected variants")

text(mp,afs+0.0003,format(round(afs,4),scientific=F))

# Clean-up
rm(nfe_mean_af,controls_mean_af,cases_mean_af,
   afs, mp)

```

### Aggregated AF

```{r}

afs <- c(nfe_aggregated_af,controls_aggregated_af,cases_aggregated_af)
names(afs) <- c("NFFE\nn=196","UBC\nn=171","CBC\nn=160")

mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("BC11 Aggregated AF: WES WECARE-NFFE\n",
                         cc_counts[1],"1KGP-NFFE,",
                         cc_counts[2],"Controls and",
                         cc_counts[3],"Cases\nCrude counts trend p=",round(trend_test$p.value,3)),  
              ylab="Aggregated AF")

text(mp,afs+0.003,format(round(afs,3),scientific=F))

# Clean-up
rm(nfe_aggregated_af,controls_aggregated_af,cases_aggregated_af,
   afs, mp, cc_counts, trend_test)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(bc11_variants.df)
dim(bc11_fiv.df)

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s03_allele_counts_bc11.RData"))

```

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
