---
title: "Ampliseq wecare only: check phenotypes"
author: "Alexey Larionov"
date: "22 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

Check phenotypes:  

- no data types should be converted  
- no misses value codes should be updated  

No data change:  
10,864 vars and 389 samples (194 UBC & 195 CBC)  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s05_update and_cleaning")
scripts_folder <- file.path(base_folder,"scripts","s05_update and_cleaning")
setwd(scripts_folder)

```

# Load data

```{r}

load(file.path(data_folder,"s01_update_genotypes.RData"))

# Update folders (just in case)
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s05_update and_cleaning")
scripts_folder <- file.path(base_folder,"scripts","s05_update and_cleaning")

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)

dim(alt_fraction.mx)

dim(gq.mx)
dim(dp.mx)

dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistence of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(alt_fraction.mx))
sum(rownames(gt_add.mx) != rownames(gq.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(alt_fraction.mx))
sum(colnames(gt_add.mx) != colnames(gq.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Check data types

```{r}

str(phenotypes.df)

```

# Check missed values

```{r}

# Check presence of NA-s
sum(is.na(phenotypes.df))

# Other possible missed value codes
sum(phenotypes.df=="",na.rm=T)
sum(phenotypes.df==".",na.rm=T)

```

# No final check and save

Nothing changed, so there is no need for final check and save

# Final section

```{r}

sessionInfo()
Sys.time()
gc()

```
