---
title: "Ampliseq wecare only: update variant annotations, apply FILTER"
author: "Alexey Larionov"
date: "23 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary  

Retain only most informative fields  
Update types of the selected fields  

Overall this drops the number of variant annotations from 116 to 65  
(may be further reduced later)  

Input data: 10,744 variants x 389 samples (194UBC + 195CBC)  
Output data: 6,839 variants x 389 samples (194UBC + 195CBC) 

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s05_update and_cleaning")
scripts_folder <- file.path(base_folder,"scripts","s05_update and_cleaning")
setwd(scripts_folder)

# Libraries
library(dplyr)

```

# Read data

```{r}

load(file.path(data_folder,"s04_explore_annotations_filter_by_hwe.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s05_update and_cleaning")
scripts_folder <- file.path(base_folder,"scripts","s05_update and_cleaning")

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)

dim(alt_fraction.mx)

dim(gq.mx)
dim(dp.mx)

dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(alt_fraction.mx))
sum(rownames(gt_add.mx) != rownames(gq.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(alt_fraction.mx))
sum(colnames(gt_add.mx) != colnames(gq.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Initial list of variant annotations

```{r}

colnames(variants.df)

```

# Keep only variants passed FILTER

```{r}

sort(table(variants.df$FILTER,useNA = "always"),decreasing=T)
pf_variants <- variants.df$FILTER == "PASS"
sum(pf_variants)

gt_add.mx <- gt_add.mx[pf_variants,]
gt_num.mx <- gt_num.mx[pf_variants,]
gt_chr.mx <- gt_chr.mx[pf_variants,]

alt_fraction.mx <- alt_fraction.mx[pf_variants,]

gq.mx <- gq.mx[pf_variants,]
dp.mx <- dp.mx[pf_variants,]

variants.df <- variants.df[pf_variants,]

rm(pf_variants)

```

# Keep selected annotations  

```{r}

selected_annottions=c(

"ID","CHROM","POS","REF","ALT",

"vep_SYMBOL","vep_VARIANT_CLASS","vep_Consequence","vep_IMPACT",
"vep_DISTANCE","vep_NEAREST","vep_STRAND",

"vep_Existing_variation","init_AC","init_AN","init_AF",
"QUAL","AS_VQSLOD","DP","MQ",
"NEGATIVE_TRAIN_SITE","POSITIVE_TRAIN_SITE",
"hwe_biallelic","InbreedingCoeff","ExcessHet",
"Multiallelic",

"vep_Allele",
"vep_Codons","vep_Amino_acids","vep_DOMAINS",
"vep_cDNA_position","vep_CCDS","vep_CDS_position","vep_Protein_position",

"SIFT_call","SIFT_score","PolyPhen_call","PolyPhen_score",

"vep_MAX_AF","vep_MAX_AF_POPS",

"vep_gnomAD_AF","vep_gnomAD_AFR_AF","vep_gnomAD_AMR_AF",
"vep_gnomAD_ASJ_AF","vep_gnomAD_EAS_AF","vep_gnomAD_FIN_AF",
"vep_gnomAD_NFE_AF","vep_gnomAD_OTH_AF","vep_gnomAD_SAS_AF",

"vep_CADD_RAW","vep_CADD_PHRED",

"CLNSIG","CLNREVSTAT","CLNDN","CLNSIGCONF","CLNVC","CLNVCSO","CLNVI",
"GENEINFO","MC","ORIGIN","RS","ALLELEID","CLNDISDB","CLNHGVS")

length(selected_annottions)
variants.df <- variants.df[,selected_annottions]
dim(variants.df)

# Clean-up
rm(selected_annottions)

```

# Update types of retained annotations in variants.df  

```{r}

str(variants.df, list.len=Inf)

variants.df$init_AC <- as.integer(variants.df$init_AC)
variants.df$init_AF <- as.numeric(variants.df$init_AF)

variants.df$AS_VQSLOD <- as.numeric(variants.df$AS_VQSLOD)

variants.df$vep_DISTANCE <- as.integer(variants.df$vep_DISTANCE)
variants.df$vep_STRAND <- as.integer(variants.df$vep_STRAND)

variants.df$vep_MAX_AF <- as.numeric(variants.df$vep_MAX_AF)

variants.df$vep_gnomAD_AF <- as.numeric(variants.df$vep_gnomAD_AF)
variants.df$vep_gnomAD_AFR_AF <- as.numeric(variants.df$vep_gnomAD_AFR_AF)
variants.df$vep_gnomAD_AMR_AF <- as.numeric(variants.df$vep_gnomAD_AMR_AF)
variants.df$vep_gnomAD_ASJ_AF <- as.numeric(variants.df$vep_gnomAD_ASJ_AF)
variants.df$vep_gnomAD_EAS_AF <- as.numeric(variants.df$vep_gnomAD_EAS_AF)
variants.df$vep_gnomAD_FIN_AF <- as.numeric(variants.df$vep_gnomAD_FIN_AF)
variants.df$vep_gnomAD_NFE_AF <- as.numeric(variants.df$vep_gnomAD_NFE_AF)
variants.df$vep_gnomAD_OTH_AF <- as.numeric(variants.df$vep_gnomAD_OTH_AF)
variants.df$vep_gnomAD_SAS_AF <- as.numeric(variants.df$vep_gnomAD_SAS_AF)

variants.df$vep_CADD_RAW <- as.numeric(variants.df$vep_CADD_RAW)
variants.df$vep_CADD_PHRED <- as.numeric(variants.df$vep_CADD_PHRED)

str(variants.df, list.len=Inf)

```

# Check data

```{r}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)

dim(alt_fraction.mx)

dim(gq.mx)
dim(dp.mx)

dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(alt_fraction.mx))
sum(rownames(gt_add.mx) != rownames(gq.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(alt_fraction.mx))
sum(colnames(gt_add.mx) != colnames(gq.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(variants.df))
sum(colnames(gt_add.mx) != rownames(phenotypes.df)) 

```

# Save result

```{r}

save.image(file.path(data_folder, "s05_update_variant_annotations_apply_FILTER.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
