---
title: "Ampliseq: bc9 OR by age"
author: "Alexey Larionov"
date: "22 Apr 2022"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Summary

Calculate CBC/UBC OR for different age groups  

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time
Sys.time()

# Memory
gc()

# Clean up
rm(list=ls())
graphics.off()

# Options
options(stringsAsFactors = F)

# Working folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","ampliseq","s01_ampliseq_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s10_age_or")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s10_age_or")
#dir.create(data_folder)

# Libraries
#library(dplyr)

```

# Load data

```{r}

load(file.path(project_folder,"data","s08_crude_counts","s03_allele_counts_bc9.RData"))

# Update folders
base_folder <- "/Users/alexey/Documents"
project_folder <- file.path(base_folder,"wecare","reanalysis_2022","ampliseq","s01_ampliseq_wecare_only")
scripts_folder <- file.path(project_folder,"scripts","s10_age_or")
setwd(scripts_folder)
data_folder <- file.path(project_folder,"data","s10_age_or")

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

table(phenotypes.df$cc, useNA = "always")

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

dim(bc9_variants.df)
dim(bc9_fiv.df)

```

# Indices

## UBC / CBC

```{r}

table(phenotypes.df$cc, useNA = "always")

ubc <- phenotypes.df$cc == 0
cbc <- phenotypes.df$cc == 1

sum(ubc)
sum(cbc)

```

## Age

```{r}

hist(phenotypes.df$sub_dx_age,
     main="Ampliseq: Ages distribution (median 51yo)",
     xlim=c(20,55))

abline(v=median(phenotypes.df$sub_dx_age), col="red", lty=2)

mean(phenotypes.df$sub_dx_age)
median(phenotypes.df$sub_dx_age)

age <- median(phenotypes.df$sub_dx_age) 
#age <- 50 # 40 & 45 do not work

young <- phenotypes.df$sub_dx_age <= age
old <- phenotypes.df$sub_dx_age > age

sum(young)
sum(old)

```

## bc9 FIVs

```{r}

bc9_fiv_ids <- bc9_fiv.df$ID
bc9_fiv_count <- nrow(bc9_fiv.df)

```

# All samples

Here the counts are calculated directly from genotypes.mx  

For all sampes, it could be extracted from bc9_fiv.df  

However, bc9_fiv.df could not be used later because  
it doesn't have age-specific counts for CBC and UBC  

## UBC ac & an

```{r}

ubc_ac <- sum(genotypes.mx[bc9_fiv_ids,ubc], na.rm=T)

ubc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids,ubc]))*2
ubc_mean_an <- ubc_total_an / bc9_fiv_count

ubc_mean_af <- ubc_ac / ubc_total_an
ubc_aggregated_af <- ubc_ac / ubc_mean_an

rm(ubc_mean_an)

```

## CBC ac & an

```{r}

cbc_ac <- sum(genotypes.mx[bc9_fiv_ids,cbc], na.rm=T)

cbc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids,cbc]))*2
cbc_mean_an <- cbc_total_an / bc9_fiv_count

cbc_mean_af <- cbc_ac / cbc_total_an
cbc_aggregated_af <- cbc_ac / cbc_mean_an

rm(cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(ubc_ac, ubc_total_an - ubc_ac,
                          cbc_ac, cbc_total_an - cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("UBC","CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(ubc_ac,ubc_total_an,
   cbc_ac,cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(ubc_mean_af,cbc_mean_af)
names(afs) <- c("UBC","CBC")

ubc_count <- sum(phenotypes.df$cc == 0)
cbc_count <- sum(phenotypes.df$cc == 1)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("bc9 Mean AF: Ampliseq all ages\n",
                ubc_count," UBC vs ",cbc_count," CBC\n",
                "Crude counts Fisher p=",round(ft$p.value,3)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(ubc_mean_af, cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(ubc_aggregated_af,cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("bc9 Aggregated AF: Ampliseq all ages\n",
                ubc_count," UBC vs ",cbc_count," CBC\n",
                "Crude counts Fisher p=",round(ft$p.value,3)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,3))

# Clean-up
rm(ubc_aggregated_af, cbc_aggregated_af, afs, mp, 
   ubc_count, cbc_count, ft)

```

# Young participants

## Young UBC ac & an

```{r}

young_ubc_ac <- sum(genotypes.mx[bc9_fiv_ids, young & ubc], na.rm=T)

young_ubc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids, young & ubc]))*2
young_ubc_mean_an <- young_ubc_total_an / bc9_fiv_count

young_ubc_mean_af <- young_ubc_ac / young_ubc_total_an
young_ubc_aggregated_af <- young_ubc_ac / young_ubc_mean_an

rm(young_ubc_mean_an)

```

## Young CBC ac & an

```{r}

young_cbc_ac <- sum(genotypes.mx[bc9_fiv_ids, young & cbc], na.rm=T)

young_cbc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids, young & cbc]))*2
young_cbc_mean_an <- young_cbc_total_an / bc9_fiv_count

young_cbc_mean_af <- young_cbc_ac / young_cbc_total_an
young_cbc_aggregated_af <- young_cbc_ac / young_cbc_mean_an

rm(young_cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(young_ubc_ac, young_ubc_total_an - young_ubc_ac,
                          young_cbc_ac, young_cbc_total_an - young_cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("Young UBC","Young CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(young_ubc_ac,young_ubc_total_an,
   young_cbc_ac,young_cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(young_ubc_mean_af,young_cbc_mean_af)
names(afs) <- c("UBC","CBC")

young_ubc_count <- sum(phenotypes.df$cc == 0 & young)
young_cbc_count <- sum(phenotypes.df$cc == 1 & young)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("bc9 Mean AF: Young Ampliseq (<=",age,")\n",
                young_ubc_count," UBC vs ",young_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(young_ubc_mean_af, young_cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(young_ubc_aggregated_af,young_cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("bc9 Aggregated AF: Young Ampliseq (<=",age,")\n",
                young_ubc_count," UBC vs ",young_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Barplot with simplified title
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("Ampliseq patients of",age,"years old or younger\n",
                young_ubc_count," UBC vs ",young_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,2),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Clean-up
rm(young_ubc_aggregated_af, young_cbc_aggregated_af, afs, mp, 
   young_ubc_count, young_cbc_count, ft)

```

# Old participants

## Old UBC ac & an

```{r}

old_ubc_ac <- sum(genotypes.mx[bc9_fiv_ids, old & ubc], na.rm=T)

old_ubc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids, old & ubc]))*2
old_ubc_mean_an <- old_ubc_total_an / bc9_fiv_count

old_ubc_mean_af <- old_ubc_ac / old_ubc_total_an
old_ubc_aggregated_af <- old_ubc_ac / old_ubc_mean_an

rm(old_ubc_mean_an)

```

## Old CBC ac & an

```{r}

old_cbc_ac <- sum(genotypes.mx[bc9_fiv_ids, old & cbc], na.rm=T)

old_cbc_total_an <- sum(!is.na(genotypes.mx[bc9_fiv_ids, old & cbc]))*2
old_cbc_mean_an <- old_cbc_total_an / bc9_fiv_count

old_cbc_mean_af <- old_cbc_ac / old_cbc_total_an
old_cbc_aggregated_af <- old_cbc_ac / old_cbc_mean_an

rm(old_cbc_mean_an)

```

## Fisher test

```{r}

fisher_matrix <- matrix(c(old_ubc_ac, old_ubc_total_an - old_ubc_ac,
                          old_cbc_ac, old_cbc_total_an - old_cbc_ac),
                          nrow=2, byrow=T)

colnames(fisher_matrix) <- c("Alt","Ref")
rownames(fisher_matrix) <- c("Old UBC","Old CBC")
fisher_matrix 

ft <- fisher.test(fisher_matrix, conf.int=T)

ft

rm(old_ubc_ac,old_ubc_total_an,
   old_cbc_ac,old_cbc_total_an,
   fisher_matrix)

```

## Barplots

### Mean AF

```{r}

# Prepare data for plot
afs <- c(old_ubc_mean_af,old_cbc_mean_af)
names(afs) <- c("UBC","CBC")

old_ubc_count <- sum(phenotypes.df$cc == 0 & old)
old_cbc_count <- sum(phenotypes.df$cc == 1 & old)

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.0005), 
              main=paste("bc9 Mean AF: Old Ampliseq (>",age,")\n",
                old_ubc_count," UBC vs ",old_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Mean AF in selected variants")

# Add labels to plot
text(mp,afs+0.0002,round(afs,4))

# Clean-up
rm(old_ubc_mean_af, old_cbc_mean_af, afs, mp)

```

### Aggregated AF

```{r}

# Prepare data for plot
afs <- c(old_ubc_aggregated_af,old_cbc_aggregated_af)
names(afs) <- c("UBC","CBC")

# Barplot (keep mid-positions of bars)
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("bc9 Aggregated AF: Old Ampliseq (>",age,")\n",
                old_ubc_count," UBC vs ",old_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Barplot with simplified title
mp <- barplot(afs, ylim=c(0,max(afs)+0.005), 
              main=paste("Ampliseq patients older than",age,"years\n",
                old_ubc_count," UBC vs ",old_cbc_count," CBC\n",
                "Crude counts Fisher p=",
                format(round(ft$p.value,4),scientific=F)),
              ylab="Aggregated AF")

# Add labels to plot
text(mp,afs+0.003,round(afs,2))

# Clean-up
rm(old_ubc_aggregated_af, old_cbc_aggregated_af, afs, mp, 
   old_ubc_count, old_cbc_count, ft,
   cbc,ubc,old,young,bc9_fiv_count,bc9_fiv_ids)

```

# Data summary

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != phenotypes.df$sample.ID)
sum(rownames(genotypes.mx) != variants.df$ID)

table(phenotypes.df$cc, useNA = "always")

dim(bc9_variants.df)
dim(bc9_fiv.df)

```

# Save results

```{r}

save.image(file.path(data_folder,"s01_age_or_bc9.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()

```
