---
title: "Ampliseq WECARE only: crude counts per variant"
author: "Alexey Larionov"
date: "24 Dec 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary

- Calculate crude counts: totals, and separately for cases and controls  
- Fisher test for the crude counts  
- FDR for all variants  

Data: 3,840 vars x 389 samples (194UBC + 195CBC)

<style>
pre{
  overflow-x: auto;
}
pre code{
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width=999)
```

# Start section

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s08_crude_counts")
#dir.create(data_folder)
scripts_folder <- file.path(base_folder,"scripts","s08_crude_counts")
setwd(scripts_folder)

# Libraries and functions
library(dplyr)
source(file.path(base_folder,"scripts","f00_functions","f01_qqunif_plot.R"))

```

# Load data

```{r}

# Load data
load(file.path(base_folder,"data","s07_relatedness_and_pca","s08_update_main_dataset.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_afterNov2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s08_crude_counts")
scripts_folder <- file.path(base_folder,"scripts","s08_crude_counts")

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Prepare indices for cases and controls

```{r}

table(phenotypes.df$cc, useNA = "always")

cases <- phenotypes.df$cc == 1
controls <- phenotypes.df$cc == 0

```

# Prepare matrix for results

```{r}

col_names <- c("id",
               "total_ac","total_an","total_af",
               "controls_ac","controls_an","controls_af",
               "cases_ac","cases_an","cases_af",
               "excess_group",
               "crude_fisher_p","crude_counts_or",
               "or_lo_ci","or_hi_ci")

variants_crude_counts.mx <- matrix(nrow=0,ncol=length(col_names))

colnames(variants_crude_counts.mx) <- col_names

rm(col_names)

```

# Calculate crude counts and Fisher test

```{r}

# For each variant
for(var in 1:nrow(variants.df)){
  
  # for testing
  # var <- 1
  
  # Get variant ID
  id <- variants.df[var,"ID"]
  
  # Get vector of genotypes
  genotypes <- genotypes.mx[var,]
  
  # Calculate crude counts
  total_ac <- sum(genotypes, na.rm=T)
  total_an <- 2*sum(!is.na(genotypes))
  total_af <- total_ac / total_an

  controls_ac <- sum(genotypes[controls], na.rm=T)
  controls_an <- 2*sum(!is.na(genotypes[controls]))
  controls_af <- controls_ac / controls_an

  cases_ac <- sum(genotypes[cases], na.rm=T)
  cases_an <- 2*sum(!is.na(genotypes[cases]))
  cases_af <- cases_ac / cases_an

  # Determine the excess group
  if(cases_af > controls_af) "cases" -> excess_group
  if(cases_af < controls_af) "controls" -> excess_group
  if(cases_af == controls_af) "none" -> excess_group
  
  # Calculate Fisher test
  fisher_matrix <- matrix(c(cases_ac, cases_an - cases_ac,
                          controls_ac, controls_an -controls_ac),
                          nrow=2, byrow=T)
  
  fisher_test <- fisher.test(fisher_matrix, conf.int=T)
  
  crude_fisher_p <- fisher_test$p.value
  crude_counts_or <- fisher_test$estimate
  crude_counts_or_ci <- fisher_test$conf.int
  
  # Compile and record the result 
  result <- c(id,
     total_ac, total_an, total_af,
     controls_ac, controls_an, controls_af,
     cases_ac, cases_an, cases_af,
     excess_group,
     crude_fisher_p, crude_counts_or, crude_counts_or_ci)
  
  variants_crude_counts.mx <- rbind(variants_crude_counts.mx,
                                    result)

  # Clean-up (to avoid chances of counts carry-over between records)
  rm(id,genotypes,
     total_ac,total_an,total_af,
     controls_ac,controls_an,controls_af,
     cases_ac,cases_an,cases_af,
     excess_group,
     fisher_matrix, fisher_test,
     crude_fisher_p, crude_counts_or, crude_counts_or_ci,
     result)
  
} # Next variant

# Check result
dim(variants_crude_counts.mx)
head(variants_crude_counts.mx)

# Clean-up
rm(var, cases, controls)

```

# Update numeric fields

```{r}

# Make data frame
variants_crude_counts.df <- as.data.frame(variants_crude_counts.mx)
dim(variants_crude_counts.df)

# Update field types
variants_crude_counts.df$id <- as.character(variants_crude_counts.df$id)

variants_crude_counts.df$total_ac <- as.integer(variants_crude_counts.df$total_ac)
variants_crude_counts.df$total_an <- as.integer(variants_crude_counts.df$total_an)
variants_crude_counts.df$total_af <- as.numeric(variants_crude_counts.df$total_af)

variants_crude_counts.df$controls_ac <- as.integer(variants_crude_counts.df$controls_ac)
variants_crude_counts.df$controls_an <- as.integer(variants_crude_counts.df$controls_an)
variants_crude_counts.df$controls_af <- as.numeric(variants_crude_counts.df$controls_af)

variants_crude_counts.df$cases_ac <- as.integer(variants_crude_counts.df$cases_ac)
variants_crude_counts.df$cases_an <- as.integer(variants_crude_counts.df$cases_an)
variants_crude_counts.df$cases_af <- as.numeric(variants_crude_counts.df$cases_af)

variants_crude_counts.df$excess_group <- as.character(variants_crude_counts.df$excess_group)

variants_crude_counts.df$crude_fisher_p <- as.numeric(variants_crude_counts.df$crude_fisher_p)
variants_crude_counts.df$crude_counts_or <- as.numeric(variants_crude_counts.df$crude_counts_or)
variants_crude_counts.df$or_lo_ci <- as.numeric(variants_crude_counts.df$or_lo_ci)
variants_crude_counts.df$or_hi_ci <- as.numeric(variants_crude_counts.df$or_hi_ci)

str(variants_crude_counts.df)

rownames(variants_crude_counts.df) <- variants_crude_counts.df$id

# Clean-up
rm(variants_crude_counts.mx)

```

# Calculate adjusted p-values 

```{r}

crude_fisher_p_adj <- p.adjust(variants_crude_counts.df$crude_fisher_p, method="BH")

variants_crude_counts.df <- cbind(variants_crude_counts.df, crude_fisher_p_adj)

rm(crude_fisher_p_adj)

```

# Add results to variants table

```{r}

sum(variants.df$ID != variants_crude_counts.df$id)

variants.df <- cbind(variants.df, variants_crude_counts.df)

variants.df <- variants.df %>% select(-id)

rm(variants_crude_counts.df)

```

# Look at the top variants

```{r}

min(variants.df$crude_fisher_p_adj)

variants.df %>% 
  select(ID,vep_SYMBOL,controls_af,cases_af,excess_group,crude_fisher_p,crude_fisher_p_adj) %>% 
  arrange(desc(crude_fisher_p)) %>% 
  filter(crude_fisher_p <= 0.01)

```

# QQ plot

```{r}

qqunif.plot(variants.df$crude_fisher_p)

rm(qqunif.plot)

```

# Check data

```{r}

ls()

# Check objects dimentions
dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)
table(phenotypes.df$cc, useNA = "always")

# Check that the tables and matrices are in sync
sum(phenotypes.df$wes_id != colnames(genotypes.mx))
sum(variants.df$ID != rownames(genotypes.mx))

```

# Save results

```{r}

save.image(file.path(data_folder,"s01_crude_counts_per_variant.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
